### 2022-12-12 bm.Rnw
coyping over from ibpf-paper/bm

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{url} 
\newcommand\eic[1]{{\textcolor{orange}{#1}}}
\newcommand\pnc[1]{{\textcolor{blue}{#1}}}
\usepackage{fullpage}
\bibliographystyle{apalike}

%% code macros
\newcommand\proglang[1]{\textsf{#1}} 
\newcommand\code[1]{\texttt{#1}}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
\newcommand\slot[1]{\code{#1}}
\newcommand\class[1]{class `\code{#1}'}
\newcommand\Class[1]{Class `\code{#1}'}

%% basic POMP definitions
\newcommand\Xspace{{\mathbb X}}
\newcommand\Yspace{{\mathbb Y}}
%\newcommand\Thetaspace{{\Theta}}
%\newcommand\vecTheta{\vec{\Theta}}
\newcommand\vecTheta{\Theta}
\newcommand\Thetaspace{\R^{\Thetadim}}
\newcommand\hatTheta{\widehat{\Theta}}
\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
\renewcommand\Xdim{D_X}
\renewcommand\Xspace{{\R^{\Xdim}}}
\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
\renewcommand\Yspace{\R}
%\newcommand\Thetadim{{\mathrm{dim}}(\Thetaspace)}
\newcommand\Thetadim{D_\theta}
\newcommand\thetadim{d_\theta}
%\newcommand\vectheta{\vec{\theta}}
\newcommand\vectheta{\theta}
\newcommand\timeSet{{\mathbb T}}
\newcommand\data[1]{#1^*}

\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
\newcommand\RP{\hspace{0.15mm}\mathrm{RP}}
\newcommand\IVP{\hspace{0.15mm}\mathrm{IVP}}

%% for bagged filters
\newcommand\ABF{ABF}
\newcommand\ABFIR{ABF-IR}
\newcommand\UBF{UBF}
\newcommand\rep{i}
\newcommand\altRep{\tilde\rep}
\newcommand\Rep{\mathcal{I}}

%% for particle filters
\newcommand\unit{u}
\newcommand\altUnit{\tilde{u}}
\newcommand\Unit{U}
\usepackage[mathscr]{euscript}
\renewcommand\time{n}
\renewcommand\vec[1]{\boldsymbol{#1}}
\newcommand\altTime{\tilde{n}}
\newcommand\Time{N}
\newcommand\Np{J}
\newcommand\np{j}
\newcommand\altNp{q}
\newcommand\altAltNp{\tilde{j}}
\newcommand\resampleIndex{r}

%% customized math macros
%\newcommand\prob[1]{\mathbb{P}\left[{#1}\right]}
\newcommand\prob{\mathbb{P}}
\newcommand\dd[1]{\mathrm{d}{#1}}
\newcommand\given{{\,\vert\,}}
\newcommand\equals{{{\,}={\,}}}
\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
\newcommand\myto{{\;:\;}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
%\newcommand\BirthDeath{\raisebox{-0.3ex}{\scalebox{1.5}{$\cdot$}}}
%\newcommand\dlta[1]{\Delta{#1}}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\normal{{\mathrm{Normal}}}
\newcommand\argequals{{\,=\,}}
\newcommand\lags{c}
\newcommand\maxlag{\overline{c}}
\newcommand\nlfList{C}
\newcommand\bigO{\mathcal{O}}
%\newcommand\loglik{\ell}
\newcommand\loglik{\lambda}
\newcommand\loglikMC{\MC{\loglik}}
\newcommand\loglikComponent{c}
\newcommand\R{\mathbb{R}}
\newcommand\param{\,;}
\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
\newcommand\MC[1]{#1^{\mbox{\tiny MC}}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}
\newcommand\Cov{\mathrm{Cov}}
\newcommand\cov{\Cov}
\newcommand\iid{\mathrm{iid}}
\newcommand\dist{\mathrm{dist}}
\newcommand\transpose{t}

<<cores,echo=F,cache=F,message = FALSE, warnings = FALSE,results='hide'>>=
library(doParallel)
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)
ggplot2::theme_set(ggplot2::theme_bw())
@

<<packages,include=F,echo=F,cache=F>>=
library("spatPomp")
library("gridExtra")
library("grid")
library("ggplot2")
library("xtable")
library("tidyverse")
library("cowplot")
library("knitr")
library("kableExtra")
stopifnot(packageVersion("pomp")>="4.1")
@

<<set-opts,include=F,cache=F,purl=F>>=

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="R> ",
        continue="+  ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )

@

<<knitr-opts,include=F,cache=F,purl=F>>=
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
#    tidy=TRUE,
#    tidy.opts=list(
#        keep.blank.line=FALSE
#    ),
    comment="",
    warning=FALSE,
    message=FALSE,
    error=TRUE,
    echo=TRUE,
    cache=FALSE,  ### SET CACHE TO TRUE WHILE EDITING TEXT
#    cache=TRUE,  ### SET CACHE TO FALSE WHILE DEBUGGING CODE
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
    fig.height=5,fig.width=8,
    out.width="\\textwidth",
    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=12
    )
)
@

<<setup,cache=F,include=F>>=
myround<- function (x, digits = 1) {
  # adapted from the broman package
  # solves the bug that round() kills significant trailing zeros
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  if (digits < 1) {
    as.character(round(x,digits))
  } else {
    tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
    zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
    tmp[tmp == paste0("-", zero)] <- zero
    tmp
  }
}
mysignif <- function (x, digits = 1) {
  myround(x, digits - ceiling(log10(abs(x))))
}
@

\title{Using an iterated block particle filter via \pkg{spatPomp}}
\author{Ning Ning and Edward L. Ionides}
\date{\today}
\begin{document}
\maketitle

The IBPF algorithm studied by \citet{ning21-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw21arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in its R Noweb (\code{.Rnw}) source file.

Consider spatial units $1,\dots,\Unit$ located evenly around a circle, where $\dist(\unit,\altUnit)$ is the circle distance,
\[
\dist(\unit,\altUnit)
= \min\big(|\unit-\altUnit|, |\unit-\altUnit+\Unit|, |\unit-\altUnit-\Unit|\big).
\]
The latent process is a $\Unit$-dimensional Brownian motion $\vec{X}(t)$ having correlation that decays with distance.
Specifically,
\[
dX_\unit(t)= \sum_{\altUnit=1}^{\Unit} \rho_{\unit}^{\dist(\unit,\altUnit)} \, dW_{\altUnit}(t),
\]
where $W_{1}(t),\dots,W_{\Unit}(t)$ are independent Brownian motions with infinitesimal variance $\sigma_{\unit}^2$, and $|\rho_{\unit}| <1$.
An observation $Y_n$ is made at each time $t_n=n$ for $n=1,2,\dots,N$, and we write $X_n=X(t_n)$.
We suppose our measurement model for discrete-time observations of the latent process is
\[
Y_{\unit,\time}=X_{\unit,\time} + \eta_{\unit,\time}
\]
where  $\eta_{\unit,\time}\overset{\text{iid}}{\sim} \normal(0,\tau_{\unit}^2)$.
The model is completed by providing the initial conditions, $\{X_\unit(0), \unit \in 1:\Unit \}$, at time $t_0=0$.
These initial conditions are specified as parameters.
An instance of this model is generated below, using the \code{bm2} function.

<<bm-setup,echo=F,eval=T>>=
set.seed(20)
@

%%%%%% iiiiiiii %%%%
<<bm-model,echo=T,eval=T>>=
library(spatPomp)
i <- 2
b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")
plot(b)
@

<<bm-followup,echo=F,eval=T>>=
U <- length(unit_names(b))
@

Here, \code{i} is a computational intensity switch which adjusts the code for varying run-time.
We set \code{i=1} for testing and debugging, and \code{i=2} for higher quality results.
For simplicity, we consider only one unit-specific parameter, $\rho_{\unit}$, with other parameters being fixed at a value shared between units.
The simulation for \code{b} has $\rho_u=\Sexpr{coef(b)["rho1"]}$ for all $\unit$, but the estimators do not know this.
Before carrying out inference, we check likelihood evaluation.
For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
This study uses a sufficiently small number of units (\code{U=\Sexpr{U}}) that the particle filter is numerically tractable.
We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.
We can readily validate the agreement between \code{bm2\_kalman\_logLik} and \code{pfilter}, and identify the likelihood cost of the block filter approximation in this situation.

<<bm_logLik_code,cache=F,echo=T,eval=F>>=
kf_logLik <- bm2_kalman_logLik(b)
pf_logLik <- replicate(10,
  logLik(pfilter(b,switch(i,10,1000)))
)
bpf_logLik2 <- replicate(10,
  logLik(bpfilter(b,switch(i,10,1000),block_size=2))
)
@

<<bm_logLik_eval,cache=F,echo=F,eval=T>>=
out_dir <- paste0("out_",i,"/")
if(!dir.exists(out_dir)) dir.create(out_dir)
stew(file=paste0(out_dir,"logLik.rda"),seed=10,{
<<bm_logLik_code>>
  bpf_logLik1 <- replicate(10,
    logLik(bpfilter(b,switch(i,10,1000),block_size=1))
  )
  bpf_logLik4 <- replicate(10,
    logLik(bpfilter(b,switch(i,10,1000),block_size=4))
  )
})
@

\begin{table}
\begin{tabular}{llrrrrr}
               & & KF & PF & BPF & BPF & BPF
	       \\
               & &  &  & ($K=1$) & ($K=2$) & ($K=4$)

\\
\hline
Log-likelihood & mean &
  \Sexpr{myround(kf_logLik,2)} &
  \Sexpr{myround(mean(pf_logLik),2)} &
  \Sexpr{myround(mean(bpf_logLik4),2)} &
  \Sexpr{myround(mean(bpf_logLik2),2)} &
  \Sexpr{myround(mean(bpf_logLik1),2)} 
\\  
               & sd &
  0.00 &
  \Sexpr{myround(sd(pf_logLik),2)} &
  \Sexpr{myround(sd(bpf_logLik4),2)} &
  \Sexpr{myround(sd(bpf_logLik2),2)} &
  \Sexpr{myround(sd(bpf_logLik1),2)} 
\end{tabular}
\caption{Likelihood evaluation for the \code{bm2} model object, \code{b}, using the Kalman filter (KF), particle filter (PF), and block particle filter (BPF) with varying numbers of blocks ($K$).
For Monte Carlo filters, the mean and standard deviation are shown for 10 replicates.
}\label{tab:lik-eval}
\end{table}

Table~\ref{tab:lik-eval} shows the increasingly negative bias, and decreasing variance, of BPF as the number of blocks increases.
For a single block, $K=1$, the BPF algorithm matches PF.
PF provides an unbiased estimate of the likelihood, and due to the convexity of the logarithm it has negative bias (approximately equal to half the variance) for estimating the log-likelihood.
Subsequently, we investigate inference for $\rho_{1:4}$ with $K=2$.

\citet{ionides21} investigated a range of values for $U$ for this model in their Figure~1, and for an epidemiological model their Figure~3.
The small scenario considered here, with $U=4$, is designed for the following purposes: (i) to validate whether or not \code{ibpf} is correctly coded by comparison with direct calculations using the Kalman filter; (ii) to check whether or not the block approximation has considerable adverse effects on inference in this case.
The inherent scalability of BPF and IBPF means that results for $U=4$ are applicable to behavior on larger systems.

For our test of IBPF, we start searches at $\rho_u=u/5$ to investigate the effect (if any) on starting value.

<<kf_mle,echo=F,eval=T>>=
stew(file=paste0(out_dir,"kf_mle.rda"),seed=999,{
  bm2_negLogLik <- function(rho){
    coef(b,names(rho)) <- unname(rho)
    -bm2_kalman_logLik(b)
  }

  rho_init <- c(rho1=0.4,rho2=0.4,rho3=0.4,rho4=0.4)
  bm2_negLogLik(rho_init)
  bm2_mle <- optim(rho_init,bm2_negLogLik)
})
@

<<kf_prof,echo=F,eval=T>>=
stew(file=paste0(out_dir,"kf_prof.rda"),seed=999,{
bm2_negProf <- function(rho_est,rho_fixed,u){
  rho_new <- rep(NA,U)
  names(rho_new) <- paste0("rho",1:U)
  rho_new[names(rho_est)] <- unname(rho_est)
  rho_new[u] <- rho_fixed
  coef(b,names(rho_new)) <- unname(rho_new)
  -bm2_kalman_logLik(b)
}

prof_init <- rep(0.4,U)
names(prof_init) <- paste0("rho",1:U)
prof_vals <- seq(from=0,to=0.8,length=switch(i,5,20))

kf_prof <- function(rho_vals,u){
  foreach(rho_prof=rho_vals)%dopar% {
    optim(prof_init[-u],bm2_negProf,rho_fixed=rho_prof,u=u)
  }
}

prof1 <- kf_prof(prof_vals,u=1)
prof2 <- kf_prof(prof_vals,u=2)
prof3 <- kf_prof(prof_vals,u=3)
prof4 <- kf_prof(prof_vals,u=4)

prof1_logLik <- sapply(prof1,function(x)-x$value)
prof2_logLik <- sapply(prof2,function(x)-x$value)
prof3_logLik <- sapply(prof3,function(x)-x$value)
prof4_logLik <- sapply(prof4,function(x)-x$value)
})

prof_range <- range(c(-bm2_mle$value,-bm2_mle$value-10))
if(0){
  par(mfrow=c(2,2))
  plot(y=prof1_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho1")
  plot(y=prof2_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho2")
  plot(y=prof3_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho3")
  plot(y=prof4_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho4")
}
@

<<ibpf-mle-code,eval=F,echo=T>>=
rho_start <- seq(from=0.2,to=0.8,length=U)
params_start <- coef(b)
params_start[paste0("rho",1:U)] <- rho_start
ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
  ibpf(b,params=params_start,
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
@

<<ibpf-mle-eval,eval=T,echo=F>>=
stew(file=paste0(out_dir,"ibpf_mle.rda"),seed=999,{
<<ibpf-mle-code>>
})

<<ibpf-mle-lik,eval=T,echo=F>>=
stew(file=paste0(out_dir,"ibpf_mle_lik.rda"),seed=888,{
ibpf_kf_eval <- sapply(ibpf_mle_searches,bm2_kalman_logLik)
ibpf_bpf_eval <- foreach(bm2fit=ibpf_mle_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
  )
}
})
@

To assess the success of these searches, we evaluate the likelihood of the resulting parameter estimates using the Kalman filter.
The highest likelihood found in these ten searches was \Sexpr{myround(max(ibpf_kf_eval),2)}
 which is not far from the actual maximum of \Sexpr{myround(-bm2_mle$value,2)}.
However, the median of \Sexpr{myround(median(ibpf_kf_eval),2)} reveals that substantial Monte Carlo maximization error is present.
It can be intractable to increase computational effort to the point where the Monte Carlo error is negligible, and instead we emphasize methods that quantify and control this.

The MLE may be of less interest than marginal confidence intervals for each unit-specific parameter.
Therefore, we compute a profile likelihood for each $u$. 
We calculate a profile likelihood for each value of $u$, using Monte Carlo adjusted profile methodology \citep{ionides17,ning21}. 
We compare this with an exact likelihood profile constructed by numerical optimization of the log-likelihood evaluated using the Kalman filter.
The IBPF implementation is identical to the search above, except that the profiled parameter is fixed.
For the profile shown in Figure~\ref{fig:ibpf-prof1-plot}, we first evaluate the likelihood using BPF rather than the Kalman filter, to present methodology applicable to non-Gaussian models.
We then check against the likelihood evaluated via the Kalman filter for the IBPF estimates, and the profile computed directly from the Kalman filter.
These reveal a distinct bias in the IBPF/BPF profile, apparently primarily to do with a bias in likelihood evaluation.
The parameter in question describes a dynamic coupling between the units, and it seems that the blocking procedure breaks some of the coupling and thereby infers a higher value of the coupling parameter that used for the simulation.
The bottom panel of Figure~\label{fig:ibpf-prof1-plot} shows that we can also diagnose this effect using the particle filter, on this small example for which the particle filter is tractable.

<<ibpf-prof,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
rho_start <- seq(from=0.2,to=0.8,length=U)
reps <- switch(i,3,10)
param_start_matrix <- matrix(params_start,byrow=TRUE,
  ncol=length(coef(b)),nrow=reps*length(prof_vals),
  dimnames=list(NULL,names(coef(b))))
param_start_matrix[,paste0("rho",u)] <- rep(prof_vals,each=reps)
ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
  ibpf(b,
    params=param_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof1time <- toc-tic
@

<<ibpf-prof-lik,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof1_lik.rda"),seed=868,{
tic <- Sys.time()
prof1_kf_eval <- sapply(ibpf_prof_searches,bm2_kalman_logLik)
ibpf_bpf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
  )
}
ibpf_pf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
  )
}
toc <- Sys.time()
})
prof1eval_time <- toc-tic
@

The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(prof1eval_time,units="mins"),2)}, shared between BPF and PF.

<<ibpf-prof1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

par(mfrow=c(3,1))
par(mai=c(0.9,0.9,0.1,0.2))

##### bpf evaluation
p1bpf <- apply(matrix(apply(ibpf_bpf_eval,2,mean),nrow=reps),2,max)
bpf_prof_range <- c(max(p1bpf)-10,max(p1bpf))
plot(y=p1bpf,x=prof_vals,ylim=bpf_prof_range,
  ylab="Log-likelihood")
p1bpf_mcap <- mcap(logLik=p1bpf,parameter=prof_vals)
lines(p1bpf_mcap$fit$parameter,p1bpf_mcap$fit$smoothed)

abline(v=p1bpf_mcap$ci)

###### kf evaluation
p1kf <- apply(matrix(prof1_kf_eval,nrow=reps),2,max)
plot(y=p1kf,x=prof_vals,ylim=prof_range,
  ylab="Log-likelihood",xlab="")
p1kf_mcap <- mcap(logLik=p1kf,parameter=prof_vals)
lines(p1kf_mcap$fit$parameter,p1kf_mcap$fit$smoothed)
p1exact_mcap <- mcap(logLik=prof1_logLik,prof_vals)
# points(y=prof1_logLik,x=prof_vals,col="red")
lines(p1exact_mcap$fit$parameter,p1exact_mcap$fit$smoothed,col="red",lty="dashed")

abline(v=p1kf_mcap$ci)
abline(v=p1exact_mcap$ci,col="red",lty="dashed")

######## pf evaluation
p1pf <- apply(matrix(apply(ibpf_pf_eval,2,mean),nrow=reps),2,max)
pf_prof_range <- c(max(p1pf)-10,max(p1pf))
plot(y=p1pf,x=prof_vals,ylim=pf_prof_range,
  ylab="Log-likelihood",xlab="rho1")
p1pf_mcap <- mcap(logLik=p1pf,parameter=prof_vals)
lines(p1pf_mcap$fit$parameter,p1pf_mcap$fit$smoothed)

abline(v=p1pf_mcap$ci)

@

We now do the same calculation for $\sigma$ for comparison with $\rho$.

<<bm-sigma-model,echo=T,eval=T>>=
set.seed(20)
b_sig <- bm2(U=4,N=switch(i,20,200),unit_specific_names="sigma")
# b_sig has the same data as b
# plot(b_sig)

  bm2_sig_negLogLik <- function(sigma){
    coef(b_sig,names(sigma)) <- unname(sigma)
    -bm2_kalman_logLik(b_sig)
  }

stew(file=paste0(out_dir,"kf_mle_sig.rda"),seed=256,{
  sigma_init <- c(sigma1=1,sigma2=1,sigma3=1,sigma4=1)
  bm2_sig_negLogLik(sigma_init)
  bm2_sig_mle <- optim(sigma_init,bm2_sig_negLogLik)
})

@

<<kf-prof-sig,echo=F,eval=T>>=
stew(file=paste0(out_dir,"kf_prof_sig.rda"),seed=512,{

bm2_sig_negProf <- function(sigma_est,sigma_fixed,u){
  sigma_new <- rep(NA,U)
  names(sigma_new) <- paste0("sigma",1:U)
  sigma_new[names(sigma_est)] <- unname(sigma_est)
  sigma_new[u] <- sigma_fixed
  coef(b_sig,names(sigma_new)) <- unname(sigma_new)
  -bm2_kalman_logLik(b_sig)
}

prof_sig_init <- rep(1,U)
names(prof_sig_init) <- paste0("sigma",1:U)
prof_sig_vals <- seq(from=0.4,to=1.6,length=switch(i,5,20))

kf_sig_prof <- function(sigma_vals,u){
  foreach(sigma_prof=sigma_vals)%dopar% {
    optim(prof_sig_init[-u],bm2_sig_negProf,sigma_fixed=sigma_prof,u=u)
  }
}

prof_sig1 <- kf_sig_prof(prof_sig_vals,u=1)
prof_sig2 <- kf_sig_prof(prof_sig_vals,u=2)
prof_sig3 <- kf_sig_prof(prof_sig_vals,u=3)
prof_sig4 <- kf_sig_prof(prof_sig_vals,u=4)

prof_sig1_logLik <- sapply(prof_sig1,function(x)-x$value)
prof_sig2_logLik <- sapply(prof_sig2,function(x)-x$value)
prof_sig3_logLik <- sapply(prof_sig3,function(x)-x$value)
prof_sig4_logLik <- sapply(prof_sig4,function(x)-x$value)
})

prof_sig_range <- range(c(-bm2_sig_mle$value,-bm2_sig_mle$value-10))
if(0){
  par(mfrow=c(2,2))
  plot(y=prof_sig1_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma1")
  plot(y=prof_sig2_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma2")
  plot(y=prof_sig3_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma3")
  plot(y=prof_sig4_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma4")
}
@

<<ibpf-prof-sig,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof_sig1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
sigma_start <- seq(from=0.4,to=1.6,length=U)
params_sig_start <- coef(b_sig)
params_sig_start[paste0("sigma",1:U)] <- sigma_start
reps <- switch(i,3,10)
param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
  ncol=length(coef(b_sig)),nrow=reps*length(prof_sig_vals),
  dimnames=list(NULL,names(coef(b_sig))))
param_sig_start_matrix[,paste0("sigma",u)] <- rep(prof_sig_vals,each=reps)
ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
  ibpf(b_sig,
    params=param_sig_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="sigma",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof_sig1time <- toc-tic
@

<<ibpf-prof-sig-lik,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof_sig1_lik.rda"),seed=868,{
tic <- Sys.time()
prof_sig1_kf_eval <- sapply(ibpf_prof_sig_searches,bm2_kalman_logLik)
ibpf_bpf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
  )
}
ibpf_pf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
  )
}
toc <- Sys.time()
})
prof_sig1eval_time <- toc-tic
@

The $\sigma_1$ profile took \Sexpr{myround(as.double(prof_sig1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(prof_sig1eval_time,units="mins"),2)}, shared between BPF and PF.

<<ibpf-prof-sig1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

par(mfrow=c(3,1))
par(mai=c(0.9,0.9,0.1,0.2))

##### bpf evaluation
p1_sig_bpf <- apply(matrix(apply(ibpf_bpf_sig_eval,2,mean),nrow=reps),2,max)
bpf_prof_sig_range <- c(max(p1_sig_bpf)-10,max(p1_sig_bpf))
plot(y=p1_sig_bpf,x=prof_sig_vals,ylim=bpf_prof_sig_range,
  ylab="Log-likelihood")
p1_sig_bpf_mcap <- mcap(logLik=p1_sig_bpf,parameter=prof_sig_vals)
lines(p1_sig_bpf_mcap$fit$parameter,p1_sig_bpf_mcap$fit$smoothed)

abline(v=p1_sig_bpf_mcap$ci)

###### kf evaluation
p1_sig_kf <- apply(matrix(prof_sig1_kf_eval,nrow=reps),2,max)
plot(y=p1_sig_kf,x=prof_sig_vals,ylim=prof_sig_range,
  ylab="Log-likelihood",xlab="")
p1_sig_kf_mcap <- mcap(logLik=p1_sig_kf,parameter=prof_sig_vals)
lines(p1_sig_kf_mcap$fit$parameter,p1_sig_kf_mcap$fit$smoothed)
p1_sig_exact_mcap <- mcap(logLik=prof_sig1_logLik,prof_sig_vals)
# points(y=prof_sig1_logLik,x=prof_sig_vals,col="red")
lines(p1_sig_exact_mcap$fit$parameter,p1_sig_exact_mcap$fit$smoothed,col="red",lty="dashed")

abline(v=p1_sig_kf_mcap$ci)
abline(v=p1_sig_exact_mcap$ci,col="red",lty="dashed")

######## pf evaluation
p1_sig_pf <- apply(matrix(apply(ibpf_pf_sig_eval,2,mean),nrow=reps),2,max)
pf_prof_sig_range <- c(max(p1_sig_pf)-10,max(p1_sig_pf))
plot(y=p1_sig_pf,x=prof_sig_vals,ylim=pf_prof_sig_range,
  ylab="Log-likelihood",xlab="sigma1")
p1_sig_pf_mcap <- mcap(logLik=p1_sig_pf,parameter=prof_sig_vals)
lines(p1_sig_pf_mcap$fit$parameter,p1_sig_pf_mcap$fit$smoothed)

abline(v=p1_sig_pf_mcap$ci)

@

We now proceed to carry out a similar analysis for the measles model generated by \code{he10}.
This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw21arxiv}.
For this model, exact likelihood evaluation is not available.
However, for a relatively small number of units ($U=4$) the particle filter provide an adequate approximation.
\citet{ionides22} considered fitting this model to data using IBPF, with 20 cities and up to $20\times 13$ parameters.
Here, our task is to focus on a smaller, simulated dataset, estimating fewer parameters in order to assess more clearly whether or not the block approximation is leading to substantial bias.

<<model-he10-code,eval=F,echo=T>>=
he10_model <- he10(U=4,dt=1/365,Tmax=1964,expandedParNames=c("R0"),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=1000,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m <- simulate(he10_model,seed=9)
@

<<model-he10-eval,eval=T,echo=F>>=
stew(file=paste0(out_dir,"m-sim.rda"),{
<<model-he10-code>>
})
plot(he10_model,log=T)
plot(m,log=T)
@

<<m-logLik-eval,cache=F,echo=F,eval=T>>=
stew(file=paste0(out_dir,"m-logLik.rda"),seed=10,{
  tic
  m_pf_logLik <- foreach(r=1:10,.combine=c) %dopar% {
    logLik(pfilter(m,switch(i,10,10000)))
  }
  m_bpf_logLik1 <- foreach(r=1:10,.combine=c) %dopar% {
    logLik(bpfilter(m,switch(i,10,10000),block_size=1))
  }
  m_bpf_logLik2 <- foreach(r=1:10,.combine=c) %dopar% {
    logLik(bpfilter(m,switch(i,10,10000),block_size=2))
  }
  m_bpf_logLik4 <- foreach(r=1:10,.combine=c) %dopar% {
    logLik(bpfilter(m,switch(i,10,10000),block_size=4))
  }
  m_lik_eval_time <- toc-tic
})

@

Likelihood evaluation took \Sexpr{myround(as.double(m_lik_eval_time,units="mins"),2)} mins.

\begin{table}
\begin{tabular}{llrrrr}
               & & PF & BPF & BPF & BPF
	       \\
               & &  & ($K=1$) & ($K=2$) & ($K=4$)

\\
\hline
Log-likelihood & mean &
  \Sexpr{myround(mean(m_pf_logLik),2)} &
  \Sexpr{myround(mean(m_bpf_logLik4),2)} &
  \Sexpr{myround(mean(m_bpf_logLik2),2)} &
  \Sexpr{myround(mean(m_bpf_logLik1),2)} 
\\  
               & sd &
  \Sexpr{myround(sd(m_pf_logLik),2)} &
  \Sexpr{myround(sd(m_bpf_logLik4),2)} &
  \Sexpr{myround(sd(m_bpf_logLik2),2)} &
  \Sexpr{myround(sd(m_bpf_logLik1),2)} 
\end{tabular}
\caption{Likelihood evaluation for the \code{he10} model object, \code{m}, using the particle filter (PF), and block particle filter (BPF) with varying numbers of blocks ($K$).
The mean and standard deviation are shown for 10 replicates with $10^4$ particles.
}\label{tab:m-lik-eval}
\end{table}

<<m-prof-sig,eval=T,echo=F>>=
stew(file=paste0(out_dir,"m_prof_R01.rda"),seed=722,{
m_prof_R0_init <- rep(30,U)
names(m_prof_R0_init) <- paste0("R0",1:U)
m_prof_R0_vals <- seq(from=15,to=45,length=switch(i,5,20))
tic <- Sys.time()
m_R0 <- m
u <- 2
R0_start <- seq(from=15,to=45,length=U)
params_R0_start <- coef(m_R0)
params_R0_start[paste0("R0",1:U)] <- R0_start
reps <- switch(i,3,10)
param_R0_start_matrix <- matrix(params_R0_start,byrow=TRUE,
  ncol=length(coef(m_R0)),nrow=reps*length(m_prof_R0_vals),
  dimnames=list(NULL,names(coef(m_R0))))
param_R0_start_matrix[,paste0("R0",u)] <- rep(m_prof_R0_vals,each=reps)
m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("R0",x))
  ibpf(m_R0,
    params=param_R0_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,20,2000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="R0",
    sharedParNames=NULL,
    block_size=1,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
m_prof_R02time <- toc-tic
@

<<m-prof-sig-lik,eval=T,echo=F>>=
stew(file=paste0(out_dir,"m_prof_R0",u,"_lik.rda"),seed=868,{
tic <- Sys.time()
m_bpf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,10),
    logLik(bpfilter(mfit,block_size=1,Np=switch(i,20,5000)))
  )
}
m_pf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,10),
    logLik(pfilter(mfit,Np=switch(i,20,5000)))
  )
}
toc <- Sys.time()
})
m_prof_R02eval_time <- toc-tic
@

The $\R0_1$ profile took \Sexpr{myround(as.double(m_prof_R02time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R02eval_time,units="mins"),2)}, shared between BPF and PF.

<<m-prof-sig1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

par(mfrow=c(2,1))
par(mai=c(0.9,0.9,0.1,0.2))

##### window to avoid problematic particle filter behavior at extremes
prof_include <- (m_prof_R0_vals > 20) & (m_prof_R0_vals < 40)

##### bpf evaluation
m_p2_R0_bpf <- apply(matrix(apply(m_bpf_R0_eval,2,mean),nrow=reps),2,max)[prof_include]
bpf_prof_R0_range <- c(max(m_p2_R0_bpf)-80,max(m_p2_R0_bpf))
plot(y=m_p2_R0_bpf,x=m_prof_R0_vals[prof_include], ylim=bpf_prof_R0_range,
  ylab="Log-likelihood")
m_p2_R0_bpf_mcap <- mcap(logLik=m_p2_R0_bpf,
  parameter=m_prof_R0_vals[prof_include])
lines(m_p2_R0_bpf_mcap$fit$parameter,m_p2_R0_bpf_mcap$fit$smoothed)

abline(v=m_p2_R0_bpf_mcap$ci)

######## pf evaluation
m_p2_R0_pf <- apply(matrix(apply(m_pf_R0_eval,2,mean),nrow=reps),2,max)[prof_include]
pf_prof_R0_range <- c(max(m_p2_R0_pf)-80,max(m_p2_R0_pf))
plot(y=m_p2_R0_pf,x=m_prof_R0_vals[prof_include], ylim=pf_prof_R0_range,
  ylab="Log-likelihood",xlab="R02")
m_p2_R0_pf_mcap <- mcap(logLik=m_p2_R0_pf,parameter=m_prof_R0_vals[prof_include])
lines(m_p2_R0_pf_mcap$fit$parameter,m_p2_R0_pf_mcap$fit$smoothed)

abline(v=m_p2_R0_pf_mcap$ci)

@

\bibliography{bib-bm}

\end{document}

### 2022-12-12 bm.tex
coyping over from ibpf-paper/bm

--- 2022-12-12 bm.Rnw
@@
-\documentclass[12pt]{article}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{xcolor}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage{amsmath,amssymb}
@@
 
-<<cores,echo=F,cache=F,message = FALSE, warnings = FALSE,results='hide'>>=
-library(doParallel)
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
-if(is.na(cores)) cores <- detectCores()
-registerDoParallel(cores)
-ggplot2::theme_set(ggplot2::theme_bw())
-@
-
-<<packages,include=F,echo=F,cache=F>>=
-library("spatPomp")
-library("gridExtra")
-library("grid")
-library("ggplot2")
-library("xtable")
-library("tidyverse")
-library("cowplot")
-library("knitr")
-library("kableExtra")
-stopifnot(packageVersion("pomp")>="4.1")
-@
-
-<<set-opts,include=F,cache=F,purl=F>>=
-
-options(
-        scipen=2,
-        help_type="html",
-        stringsAsFactors=FALSE,
-        prompt="R> ",
-        continue="+  ",
-        width=70,
-        useFancyQuotes=FALSE,
-        reindent.spaces=2,
-        xtable.comment=FALSE
-        )
-
-@
-
-<<knitr-opts,include=F,cache=F,purl=F>>=
-opts_chunk$set(
-    progress=TRUE,prompt=TRUE,highlight=FALSE,
-#    tidy=TRUE,
-#    tidy.opts=list(
-#        keep.blank.line=FALSE
-#    ),
-    comment="",
-    warning=FALSE,
-    message=FALSE,
-    error=TRUE,
-    echo=TRUE,
-    cache=FALSE,  ### SET CACHE TO TRUE WHILE EDITING TEXT
-#    cache=TRUE,  ### SET CACHE TO FALSE WHILE DEBUGGING CODE
-    strip.white=TRUE,
-    results="markup",
-    background="#FFFFFF00",
-    size="normalsize",
-    fig.path="figure/",
-    fig.lp="fig:",
-    fig.align="left",
-    fig.show="asis",
-    fig.height=5,fig.width=8,
-    out.width="\\textwidth",
-    dpi=300,
-    dev="pdf",
-    dev.args=list(
-        bg="transparent",
-        pointsize=12
-    )
-)
-@
-
-<<setup,cache=F,include=F>>=
-myround<- function (x, digits = 1) {
-  # adapted from the broman package
-  # solves the bug that round() kills significant trailing zeros
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  if (digits < 1) {
-    as.character(round(x,digits))
-  } else {
-    tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-    zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-    tmp[tmp == paste0("-", zero)] <- zero
-    tmp
-  }
-}
-mysignif <- function (x, digits = 1) {
-  myround(x, digits - ceiling(log10(abs(x))))
-}
-@
-
 \title{Using an iterated block particle filter via \pkg{spatPomp}}
@@
 \date{\today}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-<<bm-setup,echo=F,eval=T>>=
-set.seed(20)
-@
+%%%%%% iiiiiiii %%%%
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> library(spatPomp)
+R> i <- 2
+R> b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")
+R> plot(b)
+\end{verbatim}
+\end{kframe}
 
-%%%%%% iiiiiiii %%%%
-<<bm-model,echo=T,eval=T>>=
-library(spatPomp)
-i <- 2
-b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")
-plot(b)
-@
+\includegraphics[width=\textwidth]{figure/bm-model-1} \hfill{}
 
-<<bm-followup,echo=F,eval=T>>=
-U <- length(unit_names(b))
-@
+\end{knitrout}
 
@@
 For simplicity, we consider only one unit-specific parameter, $\rho_{\unit}$, with other parameters being fixed at a value shared between units.
-The simulation for \code{b} has $\rho_u=\Sexpr{coef(b)["rho1"]}$ for all $\unit$, but the estimators do not know this.
+The simulation for \code{b} has $\rho_u=0.4$ for all $\unit$, but the estimators do not know this.
 Before carrying out inference, we check likelihood evaluation.
 For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
-This study uses a sufficiently small number of units (\code{U=\Sexpr{U}}) that the particle filter is numerically tractable.
+This study uses a sufficiently small number of units (\code{U=4}) that the particle filter is numerically tractable.
 We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.
@@
 
-<<bm_logLik_code,cache=F,echo=T,eval=F>>=
-kf_logLik <- bm2_kalman_logLik(b)
-pf_logLik <- replicate(10,
-  logLik(pfilter(b,switch(i,10,1000)))
-)
-bpf_logLik2 <- replicate(10,
-  logLik(bpfilter(b,switch(i,10,1000),block_size=2))
-)
-@
-
-<<bm_logLik_eval,cache=F,echo=F,eval=T>>=
-out_dir <- paste0("out_",i,"/")
-if(!dir.exists(out_dir)) dir.create(out_dir)
-stew(file=paste0(out_dir,"logLik.rda"),seed=10,{
-<<bm_logLik_code>>
-  bpf_logLik1 <- replicate(10,
-    logLik(bpfilter(b,switch(i,10,1000),block_size=1))
-  )
-  bpf_logLik4 <- replicate(10,
-    logLik(bpfilter(b,switch(i,10,1000),block_size=4))
-  )
-})
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> kf_logLik <- bm2_kalman_logLik(b)
+R> pf_logLik <- replicate(10,
++    logLik(pfilter(b,switch(i,10,1000)))
++  )
+R> bpf_logLik2 <- replicate(10,
++    logLik(bpfilter(b,switch(i,10,1000),block_size=2))
++  )
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 
@@
 Log-likelihood & mean &
-  \Sexpr{myround(kf_logLik,2)} &
-  \Sexpr{myround(mean(pf_logLik),2)} &
-  \Sexpr{myround(mean(bpf_logLik4),2)} &
-  \Sexpr{myround(mean(bpf_logLik2),2)} &
-  \Sexpr{myround(mean(bpf_logLik1),2)} 
+  -1472.69 &
+  -1480.79 &
+  -1477.88 &
+  -1509.06 &
+  -1547.80 
 \\  
@@
   0.00 &
-  \Sexpr{myround(sd(pf_logLik),2)} &
-  \Sexpr{myround(sd(bpf_logLik4),2)} &
-  \Sexpr{myround(sd(bpf_logLik2),2)} &
-  \Sexpr{myround(sd(bpf_logLik1),2)} 
+  3.96 &
+  4.00 &
+  2.07 &
+  1.49 
 \end{tabular}
@@
 
-<<kf_mle,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"kf_mle.rda"),seed=999,{
-  bm2_negLogLik <- function(rho){
-    coef(b,names(rho)) <- unname(rho)
-    -bm2_kalman_logLik(b)
-  }
-
-  rho_init <- c(rho1=0.4,rho2=0.4,rho3=0.4,rho4=0.4)
-  bm2_negLogLik(rho_init)
-  bm2_mle <- optim(rho_init,bm2_negLogLik)
-})
-@
-
-<<kf_prof,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"kf_prof.rda"),seed=999,{
-bm2_negProf <- function(rho_est,rho_fixed,u){
-  rho_new <- rep(NA,U)
-  names(rho_new) <- paste0("rho",1:U)
-  rho_new[names(rho_est)] <- unname(rho_est)
-  rho_new[u] <- rho_fixed
-  coef(b,names(rho_new)) <- unname(rho_new)
-  -bm2_kalman_logLik(b)
-}
-
-prof_init <- rep(0.4,U)
-names(prof_init) <- paste0("rho",1:U)
-prof_vals <- seq(from=0,to=0.8,length=switch(i,5,20))
-
-kf_prof <- function(rho_vals,u){
-  foreach(rho_prof=rho_vals)%dopar% {
-    optim(prof_init[-u],bm2_negProf,rho_fixed=rho_prof,u=u)
-  }
-}
-
-prof1 <- kf_prof(prof_vals,u=1)
-prof2 <- kf_prof(prof_vals,u=2)
-prof3 <- kf_prof(prof_vals,u=3)
-prof4 <- kf_prof(prof_vals,u=4)
-
-prof1_logLik <- sapply(prof1,function(x)-x$value)
-prof2_logLik <- sapply(prof2,function(x)-x$value)
-prof3_logLik <- sapply(prof3,function(x)-x$value)
-prof4_logLik <- sapply(prof4,function(x)-x$value)
-})
-
-prof_range <- range(c(-bm2_mle$value,-bm2_mle$value-10))
-if(0){
-  par(mfrow=c(2,2))
-  plot(y=prof1_logLik,x=prof_vals,ty="l",ylim=prof_range,
-    ylab="Log-likelihood",xlab="rho1")
-  plot(y=prof2_logLik,x=prof_vals,ty="l",ylim=prof_range,
-    ylab="Log-likelihood",xlab="rho2")
-  plot(y=prof3_logLik,x=prof_vals,ty="l",ylim=prof_range,
-    ylab="Log-likelihood",xlab="rho3")
-  plot(y=prof4_logLik,x=prof_vals,ty="l",ylim=prof_range,
-    ylab="Log-likelihood",xlab="rho4")
-}
-@
-
-<<ibpf-mle-code,eval=F,echo=T>>=
-rho_start <- seq(from=0.2,to=0.8,length=U)
-params_start <- coef(b)
-params_start[paste0("rho",1:U)] <- rho_start
-ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
-  ibpf(b,params=params_start,
-    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
-    rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
-    unitParNames="rho",
-    sharedParNames=NULL,
-    block_size=2,
-    spat_regression=0,
-    cooling.fraction.50=0.5
-  )
-}
-@
-
-<<ibpf-mle-eval,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"ibpf_mle.rda"),seed=999,{
-<<ibpf-mle-code>>
-})
-
-<<ibpf-mle-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"ibpf_mle_lik.rda"),seed=888,{
-ibpf_kf_eval <- sapply(ibpf_mle_searches,bm2_kalman_logLik)
-ibpf_bpf_eval <- foreach(bm2fit=ibpf_mle_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
-  )
-}
-})
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> rho_start <- seq(from=0.2,to=0.8,length=U)
+R> params_start <- coef(b)
+R> params_start[paste0("rho",1:U)] <- rho_start
+R> ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
++    ibpf(b,params=params_start,
++      Nbpf=switch(i,2,50),Np=switch(i,20,1000),
++      rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
++      unitParNames="rho",
++      sharedParNames=NULL,
++      block_size=2,
++      spat_regression=0,
++      cooling.fraction.50=0.5
++    )
++  }
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 
 To assess the success of these searches, we evaluate the likelihood of the resulting parameter estimates using the Kalman filter.
-The highest likelihood found in these ten searches was \Sexpr{myround(max(ibpf_kf_eval),2)}
- which is not far from the actual maximum of \Sexpr{myround(-bm2_mle$value,2)}.
-However, the median of \Sexpr{myround(median(ibpf_kf_eval),2)} reveals that substantial Monte Carlo maximization error is present.
+The highest likelihood found in these ten searches was -1472.93
+ which is not far from the actual maximum of -1472.11.
+However, the median of -1475.00 reveals that substantial Monte Carlo maximization error is present.
 It can be intractable to increase computational effort to the point where the Monte Carlo error is negligible, and instead we emphasize methods that quantify and control this.
@@
 
-<<ibpf-prof,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof1.rda"),seed=722,{
-tic <- Sys.time()
-u <- 1
-rho_start <- seq(from=0.2,to=0.8,length=U)
-reps <- switch(i,3,10)
-param_start_matrix <- matrix(params_start,byrow=TRUE,
-  ncol=length(coef(b)),nrow=reps*length(prof_vals),
-  dimnames=list(NULL,names(coef(b))))
-param_start_matrix[,paste0("rho",u)] <- rep(prof_vals,each=reps)
-ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
-  rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
-  ibpf(b,
-    params=param_start_matrix[s,],
-    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
-    rw.sd=do.call(rw.sd,rw.sd_call),
-    unitParNames="rho",
-    sharedParNames=NULL,
-    block_size=2,
-    spat_regression=0,
-    cooling.fraction.50=0.5
-  )
-}
-toc <- Sys.time()
-})
-prof1time <- toc-tic
-@
+The profile took 56.11 mins using 10 computing cores.
+Likelihood evaluation took 25.07, shared between BPF and PF.
 
-<<ibpf-prof-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof1_lik.rda"),seed=868,{
-tic <- Sys.time()
-prof1_kf_eval <- sapply(ibpf_prof_searches,bm2_kalman_logLik)
-ibpf_bpf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
-  )
-}
-ibpf_pf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
-  )
-}
-toc <- Sys.time()
-})
-prof1eval_time <- toc-tic
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
 
-The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
-Likelihood evaluation took \Sexpr{myround(as.double(prof1eval_time,units="mins"),2)}, shared between BPF and PF.
+\includegraphics[width=5.5in]{figure/ibpf-prof1-plot-1} \hfill{}
 
-<<ibpf-prof1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
+\caption[Top]{Top: profile for $\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
 Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
-Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=
+Bottom: The same IBPF search with likelihood computed using the particle filter.}\label{fig:ibpf-prof1-plot}
+\end{figure}
 
-par(mfrow=c(3,1))
-par(mai=c(0.9,0.9,0.1,0.2))
-
-##### bpf evaluation
-p1bpf <- apply(matrix(apply(ibpf_bpf_eval,2,mean),nrow=reps),2,max)
-bpf_prof_range <- c(max(p1bpf)-10,max(p1bpf))
-plot(y=p1bpf,x=prof_vals,ylim=bpf_prof_range,
-  ylab="Log-likelihood")
-p1bpf_mcap <- mcap(logLik=p1bpf,parameter=prof_vals)
-lines(p1bpf_mcap$fit$parameter,p1bpf_mcap$fit$smoothed)
-
-abline(v=p1bpf_mcap$ci)
-
-###### kf evaluation
-p1kf <- apply(matrix(prof1_kf_eval,nrow=reps),2,max)
-plot(y=p1kf,x=prof_vals,ylim=prof_range,
-  ylab="Log-likelihood",xlab="")
-p1kf_mcap <- mcap(logLik=p1kf,parameter=prof_vals)
-lines(p1kf_mcap$fit$parameter,p1kf_mcap$fit$smoothed)
-p1exact_mcap <- mcap(logLik=prof1_logLik,prof_vals)
-# points(y=prof1_logLik,x=prof_vals,col="red")
-lines(p1exact_mcap$fit$parameter,p1exact_mcap$fit$smoothed,col="red",lty="dashed")
-
-abline(v=p1kf_mcap$ci)
-abline(v=p1exact_mcap$ci,col="red",lty="dashed")
-
-######## pf evaluation
-p1pf <- apply(matrix(apply(ibpf_pf_eval,2,mean),nrow=reps),2,max)
-pf_prof_range <- c(max(p1pf)-10,max(p1pf))
-plot(y=p1pf,x=prof_vals,ylim=pf_prof_range,
-  ylab="Log-likelihood",xlab="rho1")
-p1pf_mcap <- mcap(logLik=p1pf,parameter=prof_vals)
-lines(p1pf_mcap$fit$parameter,p1pf_mcap$fit$smoothed)
-
-abline(v=p1pf_mcap$ci)
-
-@
+\end{knitrout}
 
@@
 
-<<bm-sigma-model,echo=T,eval=T>>=
-set.seed(20)
-b_sig <- bm2(U=4,N=switch(i,20,200),unit_specific_names="sigma")
-# b_sig has the same data as b
-# plot(b_sig)
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> set.seed(20)
+R> b_sig <- bm2(U=4,N=switch(i,20,200),unit_specific_names="sigma")
+R> # b_sig has the same data as b
+R> # plot(b_sig)
+R> 
+R>   bm2_sig_negLogLik <- function(sigma){
++      coef(b_sig,names(sigma)) <- unname(sigma)
++      -bm2_kalman_logLik(b_sig)
++    }
+R> 
+R> stew(file=paste0(out_dir,"kf_mle_sig.rda"),seed=256,{
++    sigma_init <- c(sigma1=1,sigma2=1,sigma3=1,sigma4=1)
++    bm2_sig_negLogLik(sigma_init)
++    bm2_sig_mle <- optim(sigma_init,bm2_sig_negLogLik)
++  })
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 
-  bm2_sig_negLogLik <- function(sigma){
-    coef(b_sig,names(sigma)) <- unname(sigma)
-    -bm2_kalman_logLik(b_sig)
-  }
+The $\sigma_1$ profile took 55.87 mins using 10 computing cores.
+Likelihood evaluation took 23.72, shared between BPF and PF.
 
-stew(file=paste0(out_dir,"kf_mle_sig.rda"),seed=256,{
-  sigma_init <- c(sigma1=1,sigma2=1,sigma3=1,sigma4=1)
-  bm2_sig_negLogLik(sigma_init)
-  bm2_sig_mle <- optim(sigma_init,bm2_sig_negLogLik)
-})
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
 
-@
+\includegraphics[width=5.5in]{figure/ibpf-prof-sig1-plot-1} \hfill{}
 
-<<kf-prof-sig,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"kf_prof_sig.rda"),seed=512,{
+\caption[Top]{Top: profile for $\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
+Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
+Bottom: The same IBPF search with likelihood computed using the particle filter.}\label{fig:ibpf-prof-sig1-plot}
+\end{figure}
 
-bm2_sig_negProf <- function(sigma_est,sigma_fixed,u){
-  sigma_new <- rep(NA,U)
-  names(sigma_new) <- paste0("sigma",1:U)
-  sigma_new[names(sigma_est)] <- unname(sigma_est)
-  sigma_new[u] <- sigma_fixed
-  coef(b_sig,names(sigma_new)) <- unname(sigma_new)
-  -bm2_kalman_logLik(b_sig)
-}
-
-prof_sig_init <- rep(1,U)
-names(prof_sig_init) <- paste0("sigma",1:U)
-prof_sig_vals <- seq(from=0.4,to=1.6,length=switch(i,5,20))
-
-kf_sig_prof <- function(sigma_vals,u){
-  foreach(sigma_prof=sigma_vals)%dopar% {
-    optim(prof_sig_init[-u],bm2_sig_negProf,sigma_fixed=sigma_prof,u=u)
-  }
-}
-
-prof_sig1 <- kf_sig_prof(prof_sig_vals,u=1)
-prof_sig2 <- kf_sig_prof(prof_sig_vals,u=2)
-prof_sig3 <- kf_sig_prof(prof_sig_vals,u=3)
-prof_sig4 <- kf_sig_prof(prof_sig_vals,u=4)
-
-prof_sig1_logLik <- sapply(prof_sig1,function(x)-x$value)
-prof_sig2_logLik <- sapply(prof_sig2,function(x)-x$value)
-prof_sig3_logLik <- sapply(prof_sig3,function(x)-x$value)
-prof_sig4_logLik <- sapply(prof_sig4,function(x)-x$value)
-})
-
-prof_sig_range <- range(c(-bm2_sig_mle$value,-bm2_sig_mle$value-10))
-if(0){
-  par(mfrow=c(2,2))
-  plot(y=prof_sig1_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
-    ylab="Log-likelihood",xlab="sigma1")
-  plot(y=prof_sig2_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
-    ylab="Log-likelihood",xlab="sigma2")
-  plot(y=prof_sig3_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
-    ylab="Log-likelihood",xlab="sigma3")
-  plot(y=prof_sig4_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
-    ylab="Log-likelihood",xlab="sigma4")
-}
-@
-
-<<ibpf-prof-sig,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof_sig1.rda"),seed=722,{
-tic <- Sys.time()
-u <- 1
-sigma_start <- seq(from=0.4,to=1.6,length=U)
-params_sig_start <- coef(b_sig)
-params_sig_start[paste0("sigma",1:U)] <- sigma_start
-reps <- switch(i,3,10)
-param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
-  ncol=length(coef(b_sig)),nrow=reps*length(prof_sig_vals),
-  dimnames=list(NULL,names(coef(b_sig))))
-param_sig_start_matrix[,paste0("sigma",u)] <- rep(prof_sig_vals,each=reps)
-ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
-  rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
-  ibpf(b_sig,
-    params=param_sig_start_matrix[s,],
-    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
-    rw.sd=do.call(rw.sd,rw.sd_call),
-    unitParNames="sigma",
-    sharedParNames=NULL,
-    block_size=2,
-    spat_regression=0,
-    cooling.fraction.50=0.5
-  )
-}
-toc <- Sys.time()
-})
-prof_sig1time <- toc-tic
-@
-
-<<ibpf-prof-sig-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof_sig1_lik.rda"),seed=868,{
-tic <- Sys.time()
-prof_sig1_kf_eval <- sapply(ibpf_prof_sig_searches,bm2_kalman_logLik)
-ibpf_bpf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
-  )
-}
-ibpf_pf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
-  )
-}
-toc <- Sys.time()
-})
-prof_sig1eval_time <- toc-tic
-@
-
-The $\sigma_1$ profile took \Sexpr{myround(as.double(prof_sig1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
-Likelihood evaluation took \Sexpr{myround(as.double(prof_sig1eval_time,units="mins"),2)}, shared between BPF and PF.
-
-<<ibpf-prof-sig1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
-Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
-Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=
-
-par(mfrow=c(3,1))
-par(mai=c(0.9,0.9,0.1,0.2))
-
-##### bpf evaluation
-p1_sig_bpf <- apply(matrix(apply(ibpf_bpf_sig_eval,2,mean),nrow=reps),2,max)
-bpf_prof_sig_range <- c(max(p1_sig_bpf)-10,max(p1_sig_bpf))
-plot(y=p1_sig_bpf,x=prof_sig_vals,ylim=bpf_prof_sig_range,
-  ylab="Log-likelihood")
-p1_sig_bpf_mcap <- mcap(logLik=p1_sig_bpf,parameter=prof_sig_vals)
-lines(p1_sig_bpf_mcap$fit$parameter,p1_sig_bpf_mcap$fit$smoothed)
-
-abline(v=p1_sig_bpf_mcap$ci)
-
-###### kf evaluation
-p1_sig_kf <- apply(matrix(prof_sig1_kf_eval,nrow=reps),2,max)
-plot(y=p1_sig_kf,x=prof_sig_vals,ylim=prof_sig_range,
-  ylab="Log-likelihood",xlab="")
-p1_sig_kf_mcap <- mcap(logLik=p1_sig_kf,parameter=prof_sig_vals)
-lines(p1_sig_kf_mcap$fit$parameter,p1_sig_kf_mcap$fit$smoothed)
-p1_sig_exact_mcap <- mcap(logLik=prof_sig1_logLik,prof_sig_vals)
-# points(y=prof_sig1_logLik,x=prof_sig_vals,col="red")
-lines(p1_sig_exact_mcap$fit$parameter,p1_sig_exact_mcap$fit$smoothed,col="red",lty="dashed")
-
-abline(v=p1_sig_kf_mcap$ci)
-abline(v=p1_sig_exact_mcap$ci,col="red",lty="dashed")
-
-######## pf evaluation
-p1_sig_pf <- apply(matrix(apply(ibpf_pf_sig_eval,2,mean),nrow=reps),2,max)
-pf_prof_sig_range <- c(max(p1_sig_pf)-10,max(p1_sig_pf))
-plot(y=p1_sig_pf,x=prof_sig_vals,ylim=pf_prof_sig_range,
-  ylab="Log-likelihood",xlab="sigma1")
-p1_sig_pf_mcap <- mcap(logLik=p1_sig_pf,parameter=prof_sig_vals)
-lines(p1_sig_pf_mcap$fit$parameter,p1_sig_pf_mcap$fit$smoothed)
-
-abline(v=p1_sig_pf_mcap$ci)
-
-@
+\end{knitrout}
 
@@
 
-<<model-he10-code,eval=F,echo=T>>=
-he10_model <- he10(U=4,dt=1/365,Tmax=1964,expandedParNames=c("R0"),
-  basic_params = c(
-    alpha =0.99,      iota=0,          R0=30,
-    cohort=0.5,  amplitude=0.3,     gamma=52,
-    sigma=52,           mu=0.02,  sigmaSE=0.05,
-    rho=0.5,           psi=0.1,         g=1000,
-    S_0=0.036,         E_0=0.00007,   I_0=0.00006
-  )
-)
-m <- simulate(he10_model,seed=9)
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+R> he10_model <- he10(U=4,dt=1/365,Tmax=1964,expandedParNames=c("R0"),
++    basic_params = c(
++      alpha =0.99,      iota=0,          R0=30,
++      cohort=0.5,  amplitude=0.3,     gamma=52,
++      sigma=52,           mu=0.02,  sigmaSE=0.05,
++      rho=0.5,           psi=0.1,         g=1000,
++      S_0=0.036,         E_0=0.00007,   I_0=0.00006
++    )
++  )
+R> m <- simulate(he10_model,seed=9)
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 
-<<model-he10-eval,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"m-sim.rda"),{
-<<model-he10-code>>
-})
-plot(he10_model,log=T)
-plot(m,log=T)
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
 
-<<m-logLik-eval,cache=F,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"m-logLik.rda"),seed=10,{
-  tic
-  m_pf_logLik <- foreach(r=1:10,.combine=c) %dopar% {
-    logLik(pfilter(m,switch(i,10,10000)))
-  }
-  m_bpf_logLik1 <- foreach(r=1:10,.combine=c) %dopar% {
-    logLik(bpfilter(m,switch(i,10,10000),block_size=1))
-  }
-  m_bpf_logLik2 <- foreach(r=1:10,.combine=c) %dopar% {
-    logLik(bpfilter(m,switch(i,10,10000),block_size=2))
-  }
-  m_bpf_logLik4 <- foreach(r=1:10,.combine=c) %dopar% {
-    logLik(bpfilter(m,switch(i,10,10000),block_size=4))
-  }
-  m_lik_eval_time <- toc-tic
-})
+\includegraphics[width=\textwidth]{figure/model-he10-eval-1} \hfill{}
 
-@
+\includegraphics[width=\textwidth]{figure/model-he10-eval-2} \hfill{}
 
-Likelihood evaluation took \Sexpr{myround(as.double(m_lik_eval_time,units="mins"),2)} mins.
+\end{knitrout}
+
+Likelihood evaluation took 23.72 mins.
 
@@
 Log-likelihood & mean &
-  \Sexpr{myround(mean(m_pf_logLik),2)} &
-  \Sexpr{myround(mean(m_bpf_logLik4),2)} &
-  \Sexpr{myround(mean(m_bpf_logLik2),2)} &
-  \Sexpr{myround(mean(m_bpf_logLik1),2)} 
+  -13130.10 &
+  -13128.68 &
+  -13081.59 &
+  -13077.25 
 \\  
                & sd &
-  \Sexpr{myround(sd(m_pf_logLik),2)} &
-  \Sexpr{myround(sd(m_bpf_logLik4),2)} &
-  \Sexpr{myround(sd(m_bpf_logLik2),2)} &
-  \Sexpr{myround(sd(m_bpf_logLik1),2)} 
+  17.44 &
+  11.01 &
+  5.38 &
+  1.47 
 \end{tabular}
@@
 
-<<m-prof-sig,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"m_prof_R01.rda"),seed=722,{
-m_prof_R0_init <- rep(30,U)
-names(m_prof_R0_init) <- paste0("R0",1:U)
-m_prof_R0_vals <- seq(from=15,to=45,length=switch(i,5,20))
-tic <- Sys.time()
-m_R0 <- m
-u <- 2
-R0_start <- seq(from=15,to=45,length=U)
-params_R0_start <- coef(m_R0)
-params_R0_start[paste0("R0",1:U)] <- R0_start
-reps <- switch(i,3,10)
-param_R0_start_matrix <- matrix(params_R0_start,byrow=TRUE,
-  ncol=length(coef(m_R0)),nrow=reps*length(m_prof_R0_vals),
-  dimnames=list(NULL,names(coef(m_R0))))
-param_R0_start_matrix[,paste0("R0",u)] <- rep(m_prof_R0_vals,each=reps)
-m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
-  rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("R0",x))
-  ibpf(m_R0,
-    params=param_R0_start_matrix[s,],
-    Nbpf=switch(i,2,50),Np=switch(i,20,2000),
-    rw.sd=do.call(rw.sd,rw.sd_call),
-    unitParNames="R0",
-    sharedParNames=NULL,
-    block_size=1,
-    spat_regression=0,
-    cooling.fraction.50=0.5
-  )
-}
-toc <- Sys.time()
-})
-m_prof_R02time <- toc-tic
-@
+The $\R0_1$ profile took 1018.91 mins using 10 computing cores.
+Likelihood evaluation took 809.97, shared between BPF and PF.
 
-<<m-prof-sig-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"m_prof_R0",u,"_lik.rda"),seed=868,{
-tic <- Sys.time()
-m_bpf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,10),
-    logLik(bpfilter(mfit,block_size=1,Np=switch(i,20,5000)))
-  )
-}
-m_pf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,10),
-    logLik(pfilter(mfit,Np=switch(i,20,5000)))
-  )
-}
-toc <- Sys.time()
-})
-m_prof_R02eval_time <- toc-tic
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}
 
-The $\R0_1$ profile took \Sexpr{myround(as.double(m_prof_R02time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
-Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R02eval_time,units="mins"),2)}, shared between BPF and PF.
+\includegraphics[width=5.5in]{figure/m-prof-sig1-plot-1} \hfill{}
 
-<<m-prof-sig1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
+\caption[Top]{Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
 Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
-Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=
+Bottom: The same IBPF search with likelihood computed using the particle filter.}\label{fig:m-prof-sig1-plot}
+\end{figure}
 
-par(mfrow=c(2,1))
-par(mai=c(0.9,0.9,0.1,0.2))
-
-##### window to avoid problematic particle filter behavior at extremes
-prof_include <- (m_prof_R0_vals > 20) & (m_prof_R0_vals < 40)
-
-##### bpf evaluation
-m_p2_R0_bpf <- apply(matrix(apply(m_bpf_R0_eval,2,mean),nrow=reps),2,max)[prof_include]
-bpf_prof_R0_range <- c(max(m_p2_R0_bpf)-80,max(m_p2_R0_bpf))
-plot(y=m_p2_R0_bpf,x=m_prof_R0_vals[prof_include], ylim=bpf_prof_R0_range,
-  ylab="Log-likelihood")
-m_p2_R0_bpf_mcap <- mcap(logLik=m_p2_R0_bpf,
-  parameter=m_prof_R0_vals[prof_include])
-lines(m_p2_R0_bpf_mcap$fit$parameter,m_p2_R0_bpf_mcap$fit$smoothed)
-
-abline(v=m_p2_R0_bpf_mcap$ci)
-
-######## pf evaluation
-m_p2_R0_pf <- apply(matrix(apply(m_pf_R0_eval,2,mean),nrow=reps),2,max)[prof_include]
-pf_prof_R0_range <- c(max(m_p2_R0_pf)-80,max(m_p2_R0_pf))
-plot(y=m_p2_R0_pf,x=m_prof_R0_vals[prof_include], ylim=pf_prof_R0_range,
-  ylab="Log-likelihood",xlab="R02")
-m_p2_R0_pf_mcap <- mcap(logLik=m_p2_R0_pf,parameter=m_prof_R0_vals[prof_include])
-lines(m_p2_R0_pf_mcap$fit$parameter,m_p2_R0_pf_mcap$fit$smoothed)
-
-abline(v=m_p2_R0_pf_mcap$ci)
-
-@
+\end{knitrout}
 

### 2022-12-12 bm.Rnw
ready for testing on gl

@@
 library(spatPomp)
-i <- 2
+i <- 1
 b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")
@@
 
-<<m-prof-sig,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"m_prof_R01.rda"),seed=722,{
+<<m-prof-R0,eval=T,echo=F>>=
+stew(file=paste0(out_dir,"m_prof_R02.rda"),seed=722,{
 m_prof_R0_init <- rep(30,U)
 names(m_prof_R0_init) <- paste0("R0",1:U)
-m_prof_R0_vals <- seq(from=15,to=45,length=switch(i,5,20))
+m_prof_R0_vals <- seq(from=25,to=35,length=switch(i,5,20))
 tic <- Sys.time()
@@
 u <- 2
-R0_start <- seq(from=15,to=45,length=U)
+R0_start <- seq(from=25,to=35,length=U)
 params_R0_start <- coef(m_R0)
@@
 
-<<m-prof-sig-lik,eval=T,echo=F>>=
+<<m-prof-R0-lik,eval=T,echo=F>>=
 stew(file=paste0(out_dir,"m_prof_R0",u,"_lik.rda"),seed=868,{
@@
 
-The $\R0_1$ profile took \Sexpr{myround(as.double(m_prof_R02time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
+The $R0_2$ profile took \Sexpr{myround(as.double(m_prof_R02time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
 Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R02eval_time,units="mins"),2)}, shared between BPF and PF.
 
-<<m-prof-sig1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
-Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
+<<m-prof-R02-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_2$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
 Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=
@@
 
-##### window to avoid problematic particle filter behavior at extremes
+##### window if needed to avoid extremes with filter failures
 prof_include <- (m_prof_R0_vals > 20) & (m_prof_R0_vals < 40)
@@
 plot(y=m_p2_R0_bpf,x=m_prof_R0_vals[prof_include], ylim=bpf_prof_R0_range,
-  ylab="Log-likelihood")
+  ylab="Log-likelihood",xlab="")
 m_p2_R0_bpf_mcap <- mcap(logLik=m_p2_R0_bpf,

### 2022-12-13 bm.Rnw
m-prof-R02 results

@@
 library("spatPomp")
-library("gridExtra")
-library("grid")
+#library("gridExtra")
+#library("grid")
 library("ggplot2")
-library("xtable")
+#library("xtable")
 library("tidyverse")
-library("cowplot")
+#library("cowplot")
 library("knitr")
-library("kableExtra")
+#library("kableExtra")
 stopifnot(packageVersion("pomp")>="4.1")
@@
 library(spatPomp)
-i <- 1
+i <- 2
 b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")

### 2022-12-14 ms.Rnw
reworked ready for a recompute

--- 2022-12-12 bm.Rnw
@@
 library("spatPomp")
-library("gridExtra")
-library("grid")
 library("ggplot2")
-library("xtable")
 library("tidyverse")
-library("cowplot")
 library("knitr")
-library("kableExtra")
 stopifnot(packageVersion("pomp")>="4.1")
@@
 This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
-In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in its R Noweb (\code{.Rnw}) source file.
+We also test \code{ibpf} on simulated data for a measles transmission model.
+
+In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file.
+
+\section{Correlated Gaussian random walks} \label{sec:bm}
 
@@
 i <- 1
-b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")
+b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
 plot(b)
@@
 <<bm_logLik_eval,cache=F,echo=F,eval=T>>=
-out_dir <- paste0("out_",i,"/")
-if(!dir.exists(out_dir)) dir.create(out_dir)
-stew(file=paste0(out_dir,"logLik.rda"),seed=10,{
+bm_dir <- paste0("bm_",i,"/")
+if(!dir.exists(bm_dir)) dir.create(bm_dir)
+stew(file=paste0(bm_dir,"logLik.rda"),seed=10,{
 <<bm_logLik_code>>
@@
 <<kf_mle,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"kf_mle.rda"),seed=999,{
+stew(file=paste0(bm_dir,"kf_mle.rda"),seed=999,{
   bm2_negLogLik <- function(rho){
@@
 <<kf_prof,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"kf_prof.rda"),seed=999,{
+stew(file=paste0(bm_dir,"kf_prof.rda"),seed=999,{
 bm2_negProf <- function(rho_est,rho_fixed,u){
@@
   ibpf(b,params=params_start,
-    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
+    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
     rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
@@
 <<ibpf-mle-eval,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"ibpf_mle.rda"),seed=999,{
+stew(file=paste0(bm_dir,"ibpf_mle.rda"),seed=999,{
 <<ibpf-mle-code>>
@@
 <<ibpf-mle-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"ibpf_mle_lik.rda"),seed=888,{
+stew(file=paste0(bm_dir,"ibpf_mle_lik.rda"),seed=878,{
 ibpf_kf_eval <- sapply(ibpf_mle_searches,bm2_kalman_logLik)
 ibpf_bpf_eval <- foreach(bm2fit=ibpf_mle_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
+  replicate(switch(i,3,10),
+    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,10,1000)))
   )
@@
 <<ibpf-prof,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof1.rda"),seed=722,{
+stew(file=paste0(bm_dir,"prof1.rda"),seed=722,{
 tic <- Sys.time()
@@
 <<ibpf-prof-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof1_lik.rda"),seed=868,{
+stew(file=paste0(bm_dir,"prof1_lik.rda"),seed=868,{
 tic <- Sys.time()
@@
   replicate(switch(i,5,20),
-    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
+    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,10,1000)))
   )
@@
   replicate(switch(i,5,20),
-    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
+    logLik(pfilter(bm2fit,Np=switch(i,10,1000)))
   )
@@
 
-<<ibpf-prof1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
+<<ibpf-prof-rho-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
 Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
@@
 set.seed(20)
-b_sig <- bm2(U=4,N=switch(i,20,200),unit_specific_names="sigma")
-# b_sig has the same data as b
-# plot(b_sig)
+b_sig <- bm2(U=4,N=switch(i,10,200),unit_specific_names="sigma")
 
@@
 
-stew(file=paste0(out_dir,"kf_mle_sig.rda"),seed=256,{
+stew(file=paste0(bm_dir,"kf_mle_sig.rda"),seed=256,{
   sigma_init <- c(sigma1=1,sigma2=1,sigma3=1,sigma4=1)
@@
 <<kf-prof-sig,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"kf_prof_sig.rda"),seed=512,{
+stew(file=paste0(bm_dir,"kf_prof_sig.rda"),seed=512,{
 
@@
 <<ibpf-prof-sig,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof_sig1.rda"),seed=722,{
+stew(file=paste0(bm_dir,"prof_sig1.rda"),seed=722,{
 tic <- Sys.time()
@@
 params_sig_start[paste0("sigma",1:U)] <- sigma_start
-reps <- switch(i,3,10)
+reps <- switch(i,2,10)
 param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
@@
     params=param_sig_start_matrix[s,],
-    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
+    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
     rw.sd=do.call(rw.sd,rw.sd_call),
@@
 <<ibpf-prof-sig-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"prof_sig1_lik.rda"),seed=868,{
+stew(file=paste0(bm_dir,"prof_sig1_lik.rda"),seed=868,{
 tic <- Sys.time()
@@
 ibpf_bpf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
+  replicate(switch(i,2,10),
+    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,10,1000)))
   )
@@
 ibpf_pf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,20),
-    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
+  replicate(switch(i,2,10),
+    logLik(pfilter(bm2fit,Np=switch(i,10,1000)))
   )
@@
 
-<<ibpf-prof-sig1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
+<<ibpf-prof-sig-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units.
 Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
@@
 
+These results show that there is noticeable bias in using BPF for likelihood evaluation for this model.
+This bias leads to parameter estimation bias for IBPF.
+The same methods can be used to assess BPF and IBPF for a metapopulation model.
+We will see below that this model is well suited to BPF and IBPF.
+Intuitively, this may be because the coupling between units is weak---population movement between towns is critical to disease dynamics, but the vast majority of transmission occurs among residents of the same town.
+
+\section{A measles model} \label{sec:measles}
+
 We now proceed to carry out a similar analysis for the measles model generated by \code{he10}.
@@
 Here, our task is to focus on a smaller, simulated dataset, estimating fewer parameters in order to assess more clearly whether or not the block approximation is leading to substantial bias.
+We choose two large towns (London and Birmingham) and two small towns (Cardiff and Hastings) since we expect that population movement from large towns to small towns is essential to explain disase persistence in small towns.
+Large towns can maintain an ongoing epidemic, but, below a critical community size local extinction of the disease is expected during epidemic troughs.
 
 <<model-he10-code,eval=F,echo=T>>=
-he10_model <- he10(U=4,dt=1/365,Tmax=1964,expandedParNames=c("R0"),
+he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1964,1964),
+  expandedParNames=c("R0"),
+  towns_selected=c(1,2,11,12),
   basic_params = c(
@@
     sigma=52,           mu=0.02,  sigmaSE=0.05,
-    rho=0.5,           psi=0.1,         g=1000,
+    rho=0.5,           psi=0.1,         g=800,
     S_0=0.036,         E_0=0.00007,   I_0=0.00006
@@
 )
-m <- simulate(he10_model,seed=9)
+m_seed <- 27
+# plot(simulate(he10_model,seed=1))
+m <- simulate(he10_model,seed=m_seed)
 @
 
-<<model-he10-eval,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"m-sim.rda"),{
+<<model-he10-eval,eval=T,echo=F,fig.height=4, fig.width=6, out.width="6.5in", fig.cap = "Weekly measles case report data for four UK towns.">>=
+m_dir <- paste0("m_",i,"/")
+if(!dir.exists(m_dir)) dir.create(m_dir)
+stew(file=paste0(m_dir,"m-sim.rda"),{
 <<model-he10-code>>
 })
+par(mfrow=c(2,1))
 plot(he10_model,log=T)
+@
+
+<<he10-sim-plot,echo=F,fig.height=4, fig.width=6, out.width="6.5in", fig.cap = "Simulated data from the \\code{he10} model for four towns">>=
 plot(m,log=T)
@@
 <<m-logLik-eval,cache=F,echo=F,eval=T>>=
-stew(file=paste0(out_dir,"m-logLik.rda"),seed=10,{
+stew(file=paste0(m_dir,"m-logLik.rda"),seed=10,{
   tic
@@
   }
+  toc
   m_lik_eval_time <- toc-tic
@@
 @
-
-Likelihood evaluation took \Sexpr{myround(as.double(m_lik_eval_time,units="mins"),2)} mins.
 
@@
 
+Likelihood evaluation took \Sexpr{myround(as.double(m_lik_eval_time,units="mins"),2)} mins.
+Recall that the bias-variance trade-off for likelihood evaluation becomes a tradeoff between two sources of bias for log-likelihood evaluation, due to Jensen's inqeuality.
+Table~\ref{tab:m-lik-eval} shows that little likelihood is lost due to the block approximation for small numbers of units.
+Indeed, even for a large number of particles, a small block size gives higher log-likelihood estimates.
+This is on contrast to the results in Table~\ref{tab:lik-eval} for the spatially correlated random walk example in Section~\ref{sec:bm}.
+
+Accurate BPF likelihood evalution for a small number of units suggests that the accuracy will persist for larger numbers of units.
+It is hard to directly test this for the measles model, since we do not have an alternative accurate evalation once PF becomes inapplicable.
+However, BPF has favorable scaling properties, and if the weak coupling that makes BPF accurate on the small system is also a feature of the big system, it may be reasonable to expect accuracy in situations where it cannot be directly tested.
+
 <<m-prof-R0,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"m_prof_R02.rda"),seed=722,{
-m_prof_R0_init <- rep(30,U)
-names(m_prof_R0_init) <- paste0("R0",1:U)
+stew(file=paste0(m_dir,"m_prof_R02.rda"),seed=722,{
 m_prof_R0_vals <- seq(from=25,to=35,length=switch(i,5,20))
@@
 m_R0 <- m
-u <- 2
-R0_start <- seq(from=25,to=35,length=U)
-params_R0_start <- coef(m_R0)
-params_R0_start[paste0("R0",1:U)] <- R0_start
+m_R0_prof_u <- 1
 reps <- switch(i,3,10)
-param_R0_start_matrix <- matrix(params_R0_start,byrow=TRUE,
+param_R0_start_matrix <- matrix(coef(m_R0),byrow=TRUE,
   ncol=length(coef(m_R0)),nrow=reps*length(m_prof_R0_vals),
   dimnames=list(NULL,names(coef(m_R0))))
-param_R0_start_matrix[,paste0("R0",u)] <- rep(m_prof_R0_vals,each=reps)
+param_R0_start_matrix[,paste0("R0",m_R0_prof_u)] <- rep(m_prof_R0_vals,each=reps)
 m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
   rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("R0",x))
+  names(rw.sd_call) <- lapply((1:U)[-m_R0_prof_u],function(x)paste0("R0",x))
   ibpf(m_R0,
     params=param_R0_start_matrix[s,],
-    Nbpf=switch(i,2,50),Np=switch(i,20,2000),
+    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
     rw.sd=do.call(rw.sd,rw.sd_call),
@@
 })
-m_prof_R02time <- toc-tic
+m_prof_R0_time <- toc-tic
 @
@@
 <<m-prof-R0-lik,eval=T,echo=F>>=
-stew(file=paste0(out_dir,"m_prof_R0",u,"_lik.rda"),seed=868,{
+stew(file=paste0(m_dir,"m_prof_R0",m_R0_prof_u,"_lik.rda"),seed=868,{
 tic <- Sys.time()
 m_bpf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,10),
-    logLik(bpfilter(mfit,block_size=1,Np=switch(i,20,5000)))
+  replicate(switch(i,2,10),
+    logLik(bpfilter(mfit,block_size=1,Np=switch(i,10,5000)))
   )
@@
 m_pf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,5,10),
-    logLik(pfilter(mfit,Np=switch(i,20,5000)))
+  replicate(switch(i,2,10),
+    logLik(pfilter(mfit,Np=switch(i,10,5000)))
   )
@@
 })
-m_prof_R02eval_time <- toc-tic
+m_prof_R0_eval_time <- toc-tic
 @
 
-The $R0_2$ profile took \Sexpr{myround(as.double(m_prof_R02time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
-Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R02eval_time,units="mins"),2)}, shared between BPF and PF.
+The $R0_\Sexpr{m_R0_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_R0_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
+Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R0_eval_time,units="mins"),2)}, shared between BPF and PF.
 
-<<m-prof-R02-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_2$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
+<<m-prof-R0-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
 Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=
@@
 
-##### window if needed to avoid extremes with filter failures
-prof_include <- (m_prof_R0_vals > 20) & (m_prof_R0_vals < 40)
+##### window to avoid problematic particle filter behavior at extremes
+prof_include <- (m_prof_R0_vals > 10) & (m_prof_R0_vals < 50)
 
@@
 plot(y=m_p2_R0_pf,x=m_prof_R0_vals[prof_include], ylim=pf_prof_R0_range,
-  ylab="Log-likelihood",xlab="R02")
+  ylab="Log-likelihood",xlab="R0")
 m_p2_R0_pf_mcap <- mcap(logLik=m_p2_R0_pf,parameter=m_prof_R0_vals[prof_include])
@@
 
-\bibliography{bib-bm}
+If IBPF is effectively maximizing the BPF approximation to the likelihood, then situations where BPF has low likelihood evaluation bias may correspond to situations where IBPF has low estimation bias.
+A profile likelihood for one paramter is presented to support this, in Figure~\ref{fig:m-prof-R0-plot}.
+Here, the evaluation using PF gives a slightly tighter estimate of the profile, but this may be less accurate: PF is a higher variance algorithm, even with $U=4$, and its variance increases as the model becomes increasingly misspecified.
+Thus, the profile likelihood estimate may have additional curvature due to increasing variance (and therefore increasing Jensen bias) away from the MLE.
+
+%%%  ggggggravity parameter %%%%%%%
+
+<<model-he10-g,eval=T,echo=F>>=
+stew(file=paste0(m_dir,"m-g-sim.rda"),{
+he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1964,1964),
+  expandedParNames=c("g"),
+  towns_selected=c(1,2,11,12),
+  basic_params = c(
+    alpha =0.99,      iota=0,          R0=30,
+    cohort=0.5,  amplitude=0.3,     gamma=52,
+    sigma=52,           mu=0.02,  sigmaSE=0.05,
+    rho=0.5,           psi=0.1,         g=800,
+    S_0=0.036,         E_0=0.00007,   I_0=0.00006
+  )
+)
+m_g <- simulate(he10_model,seed=m_seed)
+if(any(obs(m_g)-obs(m_R0) != 0)) stop("m_g and m_R0 should have the same synthetic data")
+})
+@
+
+%% sssssssss
+
+<<m-slice-g,eval=T,echo=F>>=
+stew(file=paste0(m_dir,"m_slice_g.rda"),seed=722,{
+m_g_prof_u <- 4
+slice_reps <- switch(i,2,10)
+slice_points <- switch(i,5,20)
+slice_g_vals <- rep(seq(from=100,to=2500,length=slice_points),each=slice_reps)
+params_g_slice<- matrix(coef(m_g),
+  nrow=slice_reps*slice_points,
+  ncol=length(coef(m_g)),
+  byrow=T,
+  dimnames=list(NULL,names(coef(m_g))))
+params_g_slice[,paste0("g",m_g_prof_u)] <- slice_g_vals
+tic <- Sys.time()
+m_slice_g_evals <- foreach(s=1:nrow(params_g_slice),.combine=c)%dopar%{
+  logLik(bpfilter(m_g,
+    params=params_g_slice[s,],
+    Np=switch(i,10,5000),
+    block_size=1
+  ))
+}
+toc <- Sys.time()
+m_slice_g_time <- toc-tic
+})
+@
+
+<<m_slice_g_plot,eval=T,echo=F,fig.height=4, fig.width=6, out.width="5.5in", fig.cap = "Slice for $g_4$ through the true parameter vector, using BPF evaluation with $K=4$ blocks each having 1 unit.">>=
+
+plot(y=m_slice_g_evals,x=slice_g_vals,xlab="g",ylab="Log-likelihood")
+m_slice_g_mcap <- mcap(logLik=m_slice_g_evals,parameter=slice_g_vals)
+lines(m_slice_g_mcap$fit$parameter,m_slice_g_mcap$fit$smoothed)
+
+abline(v=m_slice_g_mcap$ci)
+
+@
+
+The block approximation in BPF concerns dependence between blocks and therefore may have an effect on estimation of parameters describing the coupling between units.
+The measles metapopulation model has a so-called gravity model for coupling, with a parameter $g_u$ controlling the rate of transmission from other cities into city $u$.
+A relatively straightforward way to investigate estimation of $g_u$ using BPF and IBPF is to compute a likelihood slice through the true parameter value for a simulation, with only $g_u$ being varied.
+Here, we investigate a slice for $u=4$.
+A likelihood profile cannot take a lower value than a likelihood slice, since the profile has an additional optimization.
+Therefore, a flat slice implies a flat profile; the converse is not necessarily true.
+Clear evidence of bias in a slice for a small number of units would anticipate difficulties when undertaking the more time-consuming task of obtaining a profile with many units and many parameters.
+Figure~\ref{fig:m_slice_g_plot} suggests that the BPF likelihood approximation leads to a low-bias estimate of $g_u$ when compared with Figures~\ref{fig:ibpf-prof-rho-plot} and~\ref{fig:ibpf-prof-sig-plot}.
+CHECK THAT THIS MAKES SENSE IN THE CONTEXT OF THE FINAL RESULTS.
+
+The $g_\Sexpr{m_g_prof_u}$ slice took \Sexpr{myround(as.double(m_slice_g_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
+
+<<m-prof-g,eval=F,echo=F>>=
+stew(file=paste0(m_dir,"m_prof_g.rda"),seed=722,{
+m_prof_g_vals <- seq(from=200,to=1800,length=switch(i,5,20))
+tic <- Sys.time()
+reps <- switch(i,2,10)
+param_g_start_matrix <- matrix(coef(m_g),byrow=TRUE,
+  ncol=length(coef(m_g)),nrow=reps*length(m_prof_g_vals),
+  dimnames=list(NULL,names(coef(m_g))))
+param_g_start_matrix[,paste0("g",m_g_prof_u)] <- rep(m_prof_g_vals,each=reps)
+m_prof_g_searches <- foreach(s=1:nrow(param_g_start_matrix))%dopar%{
+  rw.sd_call <- rep(list(0.02),times=U-1)
+  names(rw.sd_call) <- lapply((1:U)[-m_g_prof_u],function(x)paste0("g",x))
+  ibpf(m_g,
+    params=param_g_start_matrix[s,],
+    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
+    rw.sd=do.call(rw.sd,rw.sd_call),
+    unitParNames="g",
+    sharedParNames=NULL,
+    block_size=1,
+    spat_regression=0,
+    cooling.fraction.50=0.5
+  )
+}
+toc <- Sys.time()
+})
+m_prof_g_time <- toc-tic
+@
+
+<<m-prof-g-lik,eval=F,echo=F>>=
+stew(file=paste0(m_dir,"m_prof_g",m_g_prof_u,"_lik.rda"),seed=868,{
+tic <- Sys.time()
+m_bpf_g_eval <- foreach(mfit=m_prof_g_searches,.combine=cbind) %dopar% {
+  replicate(switch(i,2,10),
+    logLik(bpfilter(mfit,block_size=1,Np=switch(i,10,5000)))
+  )
+}
+m_pf_g_eval <- foreach(mfit=m_prof_g_searches,.combine=cbind) %dopar% {
+  replicate(switch(i,2,10),
+    logLik(pfilter(mfit,Np=switch(i,10,5000)))
+  )
+}
+toc <- Sys.time()
+})
+m_prof_g_eval_time <- toc-tic
+@
+
+%% The $g_\Sexpr{m_g_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_g_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
+%% Likelihood evaluation took \Sexpr{myround(as.double(m_prof_g_eval_time,units="mins"),2)}, shared between BPF and PF.
+
+<<m-prof-g-plot,eval=F,echo=F, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $g_4$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
+Bottom: The same IBPF search with likelihood computed using the particle filter.">>=
+
+par(mfrow=c(2,1))
+par(mai=c(0.9,0.9,0.1,0.2))
+
+##### bpf evaluation
+m_p2_g_bpf <- apply(matrix(apply(m_bpf_g_eval,2,mean),nrow=reps),2,max)
+bpf_prof_g_range <- c(max(m_p2_g_bpf)-10,max(m_p2_g_bpf))
+plot(y=m_p2_g_bpf,x=m_prof_g_vals,ylim=bpf_prof_g_range,
+  ylab="Log-likelihood",xlab="")
+m_p2_g_bpf_mcap <- mcap(logLik=m_p2_g_bpf,parameter=m_prof_g_vals)
+lines(m_p2_g_bpf_mcap$fit$parameter,m_p2_g_bpf_mcap$fit$smoothed)
+
+abline(v=m_p2_g_bpf_mcap$ci)
+
+######## pf evaluation
+m_p2_g_pf <- apply(matrix(apply(m_pf_g_eval,2,mean),nrow=reps),2,max)
+pf_prof_g_range <- c(max(m_p2_g_pf)-10,max(m_p2_g_pf))
+plot(y=m_p2_g_pf,x=m_prof_g_vals,ylim=pf_prof_g_range,
+  ylab="Log-likelihood",xlab="g")
+m_p2_g_pf_mcap <- mcap(logLik=m_p2_g_pf,parameter=m_prof_g_vals)
+lines(m_p2_g_pf_mcap$fit$parameter,m_p2_g_pf_mcap$fit$smoothed)
+
+abline(v=m_p2_g_pf_mcap$ci)
+
+@
+
+\bibliography{bib-tutorial}
 

### 2022-12-15 ms.Rnw
full recompute
tidying

@@
 library(spatPomp)
-i <- 1
+i <- 2
 b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
@@
 Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter.
-Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=
+Bottom: The same IBPF search with likelihood computed using the particle filter.
+Vertical lines show the MLE and a 95\\% confidence interval, with a dotted line at the true parameter value.",eval=T,echo=F>>=
 
@@
 abline(v=p1bpf_mcap$ci)
+abline(v=p1bpf_mcap$mle)
+abline(v=coef(b)["rho1"],lty="dotted")
 
@@
 abline(v=p1exact_mcap$ci,col="red",lty="dashed")
+abline(v=p1kf_mcap$mle)
+abline(v=coef(b)["rho1"],lty="dotted")
 
@@
 abline(v=p1pf_mcap$ci)
+abline(v=p1pf_mcap$mle)
+abline(v=coef(b)["rho1"],lty="dotted")
 
@@
 abline(v=p1_sig_bpf_mcap$ci)
+abline(v=p1_sig_bpf_mcap$mle)
+abline(v=coef(b_sig)["sigma1"],lty="dotted")
 
@@
 abline(v=p1_sig_exact_mcap$ci,col="red",lty="dashed")
+abline(v=p1_sig_kf_mcap$mle)
+abline(v=coef(b_sig)["sigma1"],lty="dotted")
 
@@
 abline(v=p1_sig_pf_mcap$ci)
+abline(v=p1_sig_pf_mcap$mle)
+abline(v=coef(b_sig)["sigma1"],lty="dotted")
 
@@
 <<m-prof-R0-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
-Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=
+Bottom: The same IBPF search with likelihood computed using the particle filter.
+Vertical lines show the MLE and a 95\\% confidence interval, with a dotted line at the true parameter value.",eval=T,echo=F>>=
 
@@
 abline(v=m_p2_R0_bpf_mcap$ci)
+abline(v=m_p2_R0_bpf_mcap$mle)
+abline(v=coef(m_R0)[paste0("R0",m_R0_prof_u)],lty="dotted")
 
@@
 abline(v=m_p2_R0_pf_mcap$ci)
+abline(v=m_p2_R0_pf_mcap$mle)
+abline(v=coef(m_R0)[paste0("R0",m_R0_prof_u)],lty="dotted")
 
@@
 
-<<m_slice_g_plot,eval=T,echo=F,fig.height=4, fig.width=6, out.width="5.5in", fig.cap = "Slice for $g_4$ through the true parameter vector, using BPF evaluation with $K=4$ blocks each having 1 unit.">>=
+<<m_slice_g_plot,eval=T,echo=F,fig.height=4, fig.width=6, out.width="5.5in", fig.cap = "Slice for $g_4$ through the true parameter vector, using BPF evaluation with $K=4$ blocks each having 1 unit.
+Vertical lines show the MLE and a 95\\% confidence interval, with a dotted line at the true parameter value.">>=
 
@@
 abline(v=m_slice_g_mcap$ci)
+abline(v=m_slice_g_mcap$mle)
+abline(v=coef(m_g)[paste0("g",m_g_prof_u)],lty="dotted")
 
@@
 Clear evidence of bias in a slice for a small number of units would anticipate difficulties when undertaking the more time-consuming task of obtaining a profile with many units and many parameters.
-Figure~\ref{fig:m_slice_g_plot} suggests that the BPF likelihood approximation leads to a low-bias estimate of $g_u$ when compared with Figures~\ref{fig:ibpf-prof-rho-plot} and~\ref{fig:ibpf-prof-sig-plot}.
-CHECK THAT THIS MAKES SENSE IN THE CONTEXT OF THE FINAL RESULTS.
+Figure~\ref{fig:m_slice_g_plot} suggests that the BPF likelihood approximation leads to a lower-bias estimate of $g_u$ than for the coupling parameter $\rho$ in the Gaussian model (Figure~\ref{fig:ibpf-prof-rho-plot}).
+Figure~\ref{fig:m_slice_g_plot} is consistent with a small positive bias: BPF ignores part of the dependence in the filter distribution, so it may be expected that it compensates by a bias toward parameter values that increase the coupling.
 

### 2022-12-20 ms.Rnw
revised ready for recompute

@@
 %% customized math macros
-%\newcommand\prob[1]{\mathbb{P}\left[{#1}\right]}
 \newcommand\prob{\mathbb{P}}
@@
 \newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-%\newcommand\BirthDeath{\raisebox{-0.3ex}{\scalebox{1.5}{$\cdot$}}}
-%\newcommand\dlta[1]{\Delta{#1}}
 \newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
@@
 \newcommand\bigO{\mathcal{O}}
-%\newcommand\loglik{\ell}
 \newcommand\loglik{\lambda}
@@
 We also test \code{ibpf} on simulated data for a measles transmission model.
-
 In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file.
@@
 library(spatPomp)
-i <- 2
+i <- 1
 b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
@@
     block_size=2,
-    spat_regression=0,
     cooling.fraction.50=0.5
@@
 The IBPF implementation is identical to the search above, except that the profiled parameter is fixed.
-For the profile shown in Figure~\ref{fig:ibpf-prof1-plot}, we first evaluate the likelihood using BPF rather than the Kalman filter, to present methodology applicable to non-Gaussian models.
+For the profile shown in Figure~\ref{fig:ibpf-prof-rho-plot}, we first evaluate the likelihood using BPF rather than the Kalman filter, to present methodology applicable to non-Gaussian models.
 We then check against the likelihood evaluated via the Kalman filter for the IBPF estimates, and the profile computed directly from the Kalman filter.
@@
 The parameter in question describes a dynamic coupling between the units, and it seems that the blocking procedure breaks some of the coupling and thereby infers a higher value of the coupling parameter that used for the simulation.
-The bottom panel of Figure~\label{fig:ibpf-prof1-plot} shows that we can also diagnose this effect using the particle filter, on this small example for which the particle filter is tractable.
+The bottom panel of Figure~\label{fig:ibpf-prof-rho-plot} shows that we can also diagnose this effect using the particle filter, on this small example for which the particle filter is tractable.
 
@@
 toc <- Sys.time()
+prof1cores <- cores
 })
@@
 
-The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
+The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{prof1cores} computing cores.
 Likelihood evaluation took \Sexpr{myround(as.double(prof1eval_time,units="mins"),2)}, shared between BPF and PF.
@@
 plot(y=p1bpf,x=prof_vals,ylim=bpf_prof_range,
-  ylab="Log-likelihood")
+  ylab="Log-likelihood",xlab="")
 p1bpf_mcap <- mcap(logLik=p1bpf,parameter=prof_vals)
@@
 
-We now do the same calculation for $\sigma$ for comparison with $\rho$.
-
-<<bm-sigma-model,echo=T,eval=T>>=
+<<bm-sigma-model,echo=F,eval=T>>=
 set.seed(20)
@@
 toc <- Sys.time()
+prof_sig_cores <- cores
 })
@@
 @
-
-The $\sigma_1$ profile took \Sexpr{myround(as.double(prof_sig1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
-Likelihood evaluation took \Sexpr{myround(as.double(prof_sig1eval_time,units="mins"),2)}, shared between BPF and PF.
 
@@
 plot(y=p1_sig_bpf,x=prof_sig_vals,ylim=bpf_prof_sig_range,
-  ylab="Log-likelihood")
+  ylab="Log-likelihood",xlab="")
 p1_sig_bpf_mcap <- mcap(logLik=p1_sig_bpf,parameter=prof_sig_vals)
@@
 
-These results show that there is noticeable bias in using BPF for likelihood evaluation for this model.
-This bias leads to parameter estimation bias for IBPF.
+We now do the same calculation for $\sigma$, shown in Figure~\ref{fig:ibpf-prof-sig-plot},  for comparison with $\rho$.
+For this model, we see that there is noticeable discrepancy between the true profile and that obtained using BPF for likelihood evaluation.
+This bias is stronger for $\rho$  than $\sigma$, and may lead to parameter estimation bias for IBPF.
 The same methods can be used to assess BPF and IBPF for a metapopulation model.
@@
 <<model-he10-code,eval=F,echo=T>>=
-he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1964,1964),
+he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1955,1964),
   expandedParNames=c("R0"),
@@
 )
-m_seed <- 27
-# plot(simulate(he10_model,seed=1))
-m <- simulate(he10_model,seed=m_seed)
+m <- simulate(he10_model,seed=27)
 @
@@
     block_size=1,
-    spat_regression=0,
     cooling.fraction.50=0.5
@@
 m_prof_R0_eval_time <- toc-tic
+prof_R0_cores <- cores
 @
 
-The $R0_\Sexpr{m_R0_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_R0_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
+The $R0_\Sexpr{m_R0_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_R0_time,units="mins"),2)} mins using \Sexpr{prof_R0_cores} computing cores.
 Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R0_eval_time,units="mins"),2)}, shared between BPF and PF.
@@
 stew(file=paste0(m_dir,"m-g-sim.rda"),{
-he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1964,1964),
+he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1955,1964),
   expandedParNames=c("g"),
@@
 )
-m_g <- simulate(he10_model,seed=m_seed)
-if(any(obs(m_g)-obs(m_R0) != 0)) stop("m_g and m_R0 should have the same synthetic data")
+m_g <- simulate(he10_model,seed=27)
+if(dim(obs(m_g))!=dim(obs(m_R0)) || any(obs(m_g)-obs(m_R0) != 0)) stop("m_g and m_R0 should have the same synthetic data")
 })
@@
 m_slice_g_time <- toc-tic
+m_slice_cores <- cores
 })
@@
 
-The $g_\Sexpr{m_g_prof_u}$ slice took \Sexpr{myround(as.double(m_slice_g_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
-
-<<m-prof-g,eval=F,echo=F>>=
-stew(file=paste0(m_dir,"m_prof_g.rda"),seed=722,{
-m_prof_g_vals <- seq(from=200,to=1800,length=switch(i,5,20))
-tic <- Sys.time()
-reps <- switch(i,2,10)
-param_g_start_matrix <- matrix(coef(m_g),byrow=TRUE,
-  ncol=length(coef(m_g)),nrow=reps*length(m_prof_g_vals),
-  dimnames=list(NULL,names(coef(m_g))))
-param_g_start_matrix[,paste0("g",m_g_prof_u)] <- rep(m_prof_g_vals,each=reps)
-m_prof_g_searches <- foreach(s=1:nrow(param_g_start_matrix))%dopar%{
-  rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-m_g_prof_u],function(x)paste0("g",x))
-  ibpf(m_g,
-    params=param_g_start_matrix[s,],
-    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
-    rw.sd=do.call(rw.sd,rw.sd_call),
-    unitParNames="g",
-    sharedParNames=NULL,
-    block_size=1,
-    spat_regression=0,
-    cooling.fraction.50=0.5
-  )
-}
-toc <- Sys.time()
-})
-m_prof_g_time <- toc-tic
-@
-
-<<m-prof-g-lik,eval=F,echo=F>>=
-stew(file=paste0(m_dir,"m_prof_g",m_g_prof_u,"_lik.rda"),seed=868,{
-tic <- Sys.time()
-m_bpf_g_eval <- foreach(mfit=m_prof_g_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,2,10),
-    logLik(bpfilter(mfit,block_size=1,Np=switch(i,10,5000)))
-  )
-}
-m_pf_g_eval <- foreach(mfit=m_prof_g_searches,.combine=cbind) %dopar% {
-  replicate(switch(i,2,10),
-    logLik(pfilter(mfit,Np=switch(i,10,5000)))
-  )
-}
-toc <- Sys.time()
-})
-m_prof_g_eval_time <- toc-tic
-@
-
-%% The $g_\Sexpr{m_g_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_g_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
-%% Likelihood evaluation took \Sexpr{myround(as.double(m_prof_g_eval_time,units="mins"),2)}, shared between BPF and PF.
-
-<<m-prof-g-plot,eval=F,echo=F, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $g_4$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit.
-Bottom: The same IBPF search with likelihood computed using the particle filter.">>=
-
-par(mfrow=c(2,1))
-par(mai=c(0.9,0.9,0.1,0.2))
-
-##### bpf evaluation
-m_p2_g_bpf <- apply(matrix(apply(m_bpf_g_eval,2,mean),nrow=reps),2,max)
-bpf_prof_g_range <- c(max(m_p2_g_bpf)-10,max(m_p2_g_bpf))
-plot(y=m_p2_g_bpf,x=m_prof_g_vals,ylim=bpf_prof_g_range,
-  ylab="Log-likelihood",xlab="")
-m_p2_g_bpf_mcap <- mcap(logLik=m_p2_g_bpf,parameter=m_prof_g_vals)
-lines(m_p2_g_bpf_mcap$fit$parameter,m_p2_g_bpf_mcap$fit$smoothed)
-
-abline(v=m_p2_g_bpf_mcap$ci)
-
-######## pf evaluation
-m_p2_g_pf <- apply(matrix(apply(m_pf_g_eval,2,mean),nrow=reps),2,max)
-pf_prof_g_range <- c(max(m_p2_g_pf)-10,max(m_p2_g_pf))
-plot(y=m_p2_g_pf,x=m_prof_g_vals,ylim=pf_prof_g_range,
-  ylab="Log-likelihood",xlab="g")
-m_p2_g_pf_mcap <- mcap(logLik=m_p2_g_pf,parameter=m_prof_g_vals)
-lines(m_p2_g_pf_mcap$fit$parameter,m_p2_g_pf_mcap$fit$smoothed)
-
-abline(v=m_p2_g_pf_mcap$ci)
-
-@
+The $g_\Sexpr{m_g_prof_u}$ slice took \Sexpr{myround(as.double(m_slice_g_time,units="mins"),2)} mins using \Sexpr{m_slice_cores} computing cores.
 

### 2022-12-21 ms.Rnw
recomputed
edits

@@
 The small scenario considered here, with $U=4$, is designed for the following purposes: (i) to validate whether or not \code{ibpf} is correctly coded by comparison with direct calculations using the Kalman filter; (ii) to check whether or not the block approximation has considerable adverse effects on inference in this case.
-The inherent scalability of BPF and IBPF means that results for $U=4$ are applicable to behavior on larger systems.
+The inherent scalability of BPF and IBPF, together with the spatial homogeneity of the model, suggests that results for $U=4$ should be representative for larger $U$.
 
@@
 However, the median of \Sexpr{myround(median(ibpf_kf_eval),2)} reveals that substantial Monte Carlo maximization error is present.
-It can be intractable to increase computational effort to the point where the Monte Carlo error is negligible, and instead we emphasize methods that quantify and control this.
+On harder problems, it can be intractable to increase computational effort to the point where the Monte Carlo error is negligible, and instead we emphasize methods that quantify and control the error.
 
 The MLE may be of less interest than marginal confidence intervals for each unit-specific parameter.
-Therefore, we compute a profile likelihood for each $u$. 
-We calculate a profile likelihood for each value of $u$, using Monte Carlo adjusted profile methodology \citep{ionides17,ning21}. 
+Therefore, we compute a profile likelihood for $\rho_u$, using Monte Carlo adjusted profile methodology \citep{ionides17,ning21}. 
 We compare this with an exact likelihood profile constructed by numerical optimization of the log-likelihood evaluated using the Kalman filter.
@@
 These reveal a distinct bias in the IBPF/BPF profile, apparently primarily to do with a bias in likelihood evaluation.
-The parameter in question describes a dynamic coupling between the units, and it seems that the blocking procedure breaks some of the coupling and thereby infers a higher value of the coupling parameter that used for the simulation.
+The parameter in question describes a dynamic coupling between the units, and a heuristic explanation may be that the blocking procedure breaks some of the coupling and thereby leads to a higher inferred value of the coupling parameter to counterbalance that bias.
 The bottom panel of Figure~\label{fig:ibpf-prof-rho-plot} shows that we can also diagnose this effect using the particle filter, on this small example for which the particle filter is tractable.
@@
 
-The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{prof1cores} computing cores.
-Likelihood evaluation took \Sexpr{myround(as.double(prof1eval_time,units="mins"),2)}, shared between BPF and PF.
+The profile maximization took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{prof1cores} computing cores.
 
@@
 We now do the same calculation for $\sigma$, shown in Figure~\ref{fig:ibpf-prof-sig-plot},  for comparison with $\rho$.
-For this model, we see that there is noticeable discrepancy between the true profile and that obtained using BPF for likelihood evaluation.
-This bias is stronger for $\rho$  than $\sigma$, and may lead to parameter estimation bias for IBPF.
-The same methods can be used to assess BPF and IBPF for a metapopulation model.
+There is some evidence of a bias for $\sigma$, but the issue appears to be more severe for $\rho$.
+
+In Figures~\ref{fig:ibpf-prof-sig-plot} and~\ref{fig:ibpf-prof-rho-plot}, we see that the particle filter provides a close approximation to the exact likelihood, though even for 4 units the variablility is noticeable.
+Comparison with PF for a small number of units can be used to assess BPF and IBPF for a metapopulation model.
 We will see below that this model is well suited to BPF and IBPF.
@@
 
-We now proceed to carry out a similar analysis for the measles model generated by \code{he10}.
+We now proceed to carry out a similar analysis for the measles model generated by the \pkg{spatPomp} function \code{he10}.
 This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw21arxiv}.
@@
 toc <- Sys.time()
+prof_R0_cores <- cores
 })
@@
 toc <- Sys.time()
+prof_R0_eval_cores <- cores
 })
 m_prof_R0_eval_time <- toc-tic
-prof_R0_cores <- cores
 @
@@
 The $R0_\Sexpr{m_R0_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_R0_time,units="mins"),2)} mins using \Sexpr{prof_R0_cores} computing cores.
-Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R0_eval_time,units="mins"),2)}, shared between BPF and PF.
+Subsequent likelihood evaluation at the proposed profile points took \Sexpr{myround(as.double(m_prof_R0_eval_time,units="mins"),2)} mins, for BPF and PF combined.
+Accurate evaluation of the likelihood is important for inference, and optimization has been empirically found to require fewer particles than evaluation.
+This is a large computing time for one profile with only 4 units.
+Perhaps satisfactory results could have been generated more quickly, but generally one expects that simulation based inference for highly nonlinear and non-Gaussian spatiotemporal systems is going to require a considerable amount of computation.
 
@@
 par(mfrow=c(2,1))
-par(mai=c(0.9,0.9,0.1,0.2))
+par(mai=c(0.9,0.9,0.15,0.2))
 
@@
 Clear evidence of bias in a slice for a small number of units would anticipate difficulties when undertaking the more time-consuming task of obtaining a profile with many units and many parameters.
-Figure~\ref{fig:m_slice_g_plot} suggests that the BPF likelihood approximation leads to a lower-bias estimate of $g_u$ than for the coupling parameter $\rho$ in the Gaussian model (Figure~\ref{fig:ibpf-prof-rho-plot}).
 Figure~\ref{fig:m_slice_g_plot} is consistent with a small positive bias: BPF ignores part of the dependence in the filter distribution, so it may be expected that it compensates by a bias toward parameter values that increase the coupling.
 
-The $g_\Sexpr{m_g_prof_u}$ slice took \Sexpr{myround(as.double(m_slice_g_time,units="mins"),2)} mins using \Sexpr{m_slice_cores} computing cores.
+A slice is equivalent to a profile with only a single estimated parameter, so we can construct a Monte Carlo adjusted profile confidence interval under this assumption.
+The $g_\Sexpr{m_g_prof_u}$ slice took \Sexpr{myround(as.double(m_slice_g_time,units="mins"),2)} mins using \Sexpr{m_slice_cores} computing cores, which was considerably less computational effort than was used for the profile.
+Typically, sliced likelihood plots are a computationally convenient tool used for preliminary investigations, and a full profile is preferred for a final conclusion.
 

### 2022-12-22 ms.Rnw
a few edits

@@
 
+\clearpage
+
 \section{A measles model} \label{sec:measles}
@@
 
-<<model-he10-eval,eval=T,echo=F,fig.height=4, fig.width=6, out.width="6.5in", fig.cap = "Weekly measles case report data for four UK towns.">>=
+<<model-he10-plot,eval=T,echo=F,fig.height=7, fig.width=6, out.width="6.5in", fig.cap = "Weekly measles case reports. (A) Data for four UK towns. (B) Simulated data.">>=
 m_dir <- paste0("m_",i,"/")
@@
 })
-par(mfrow=c(2,1))
-plot(he10_model,log=T)
-@
-
-<<he10-sim-plot,echo=F,fig.height=4, fig.width=6, out.width="6.5in", fig.cap = "Simulated data from the \\code{he10} model for four towns">>=
-plot(m,log=T)
+library(grid)
+library(cowplot)
+plot_data <- plot(he10_model,log=T)+xlab("")
+plot_sim <- plot(m,log=T)+xlab("Year")
+plot_grid(
+  plot_data,
+  plot_sim,
+  labels=c("A","B"),
+  nrow=2,ncol=1
+)
 @

### 2023-04-27 ms.Rnw
minor updates, without re-running code

@@
 \author{Ning Ning and Edward L. Ionides}
-\date{\today}
+\date{\small Compiled \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}\\
+Source code at \url{https://github.com/ionides/ibpf-tutorial}}
 \begin{document}
@@
 
-The IBPF algorithm studied by \citet{ning21-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw21arxiv} as the function  \code{ibpf}.
+The IBPF algorithm studied by \citet{ning23-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw23arxiv} as the function  \code{ibpf}.
 This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
@@
 We now proceed to carry out a similar analysis for the measles model generated by the \pkg{spatPomp} function \code{he10}.
-This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw21arxiv}.
+This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw23arxiv}.
 For this model, exact likelihood evaluation is not available.

### 2023-05-01 ms.Rnw
code updated for pomp 5

@@
 stew(file=paste0(bm_dir,"logLik.rda"),seed=10,{
+
+  cat(capture.output(sessionInfo()),
+    file=paste0(bm_dir,"sessionInfo.txt"),sep="\n")
+
 <<bm_logLik_code>>
@@
     Nbpf=switch(i,2,50),Np=switch(i,10,1000),
-    rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
+    rw.sd=rw_sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
     unitParNames="rho",
@@
   ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
-  rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
+    rw_sd_call <- rep(list(0.02),times=U-1)
+    names(rw_sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
     ibpf(b,
@@
       Nbpf=switch(i,2,50),Np=switch(i,20,1000),
-    rw.sd=do.call(rw.sd,rw.sd_call),
+      rw.sd=do.call(rw_sd,rw_sd_call),
       unitParNames="rho",
@@
 stew(file=paste0(m_dir,"m-sim.rda"),{
+  cat(capture.output(sessionInfo()),
+    file=paste0(m_dir,"sessionInfo.txt"),sep="\n")
 <<model-he10-code>>
@@
 m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
-  rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-m_R0_prof_u],function(x)paste0("R0",x))
+  rw_sd_call <- rep(list(0.02),times=U-1)
+  names(rw_sd_call) <- lapply((1:U)[-m_R0_prof_u],function(x)paste0("R0",x))
   ibpf(m_R0,
@@
     Nbpf=switch(i,2,50),Np=switch(i,10,2000),
-    rw.sd=do.call(rw.sd,rw.sd_call),
+    rw.sd=do.call(rw_sd,rw_sd_call),
     unitParNames="R0",

### 2023-05-02 ms.Rnw
recomputed
debugged

@@
 ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
-  rw.sd_call <- rep(list(0.02),times=U-1)
-  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
+  rw_sd_call <- rep(list(0.02),times=U-1)
+  names(rw_sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
   ibpf(b_sig,
@@
     Nbpf=switch(i,2,50),Np=switch(i,10,1000),
-    rw.sd=do.call(rw.sd,rw.sd_call),
+    rw.sd=do.call(rw_sd,rw_sd_call),
     unitParNames="sigma",

### 2023-06-05 ms.Rnw
diagnostics section added

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage{amsmath,amssymb}
@@
 
+\begin{abstract}
 The IBPF algorithm studied by \citet{ning23-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw23arxiv} as the function  \code{ibpf}.
@@
 In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file. 
-
-\section{Correlated Gaussian random walks} \label{sec:bm}
+\end{abstract}
+
+\tableofcontents
+
+\newpage
+
+\section{A toy example: Correlated Gaussian random walks} \label{sec:bm}
 
@@
 
+\section{Diagnostics}
+
+<<diagnostics-setup,eval=T,echo=F>>=
+diagnostics_dir <- paste0("diagnostics_",i,"/")
+if(!dir.exists(diagnostics_dir)) dir.create(diagnostics_dir)
+@
+
+If likelihood maximization appears to be failing, or has unacceptably high Monte Carlo variability, one wishes to diagnose the cause, or causes.
+Even if inference appears to be operating satisfactorily, it is good practice to see whether there are some data points for which the model is inappropriate, or some parameters for which optimization is less stable.
+It may be interesting to know if less Monte Carlo effort would be sufficient, or if more would be useful.
+For all these questions, a starting point is exploratory analysis of diagnostic data produced by \code{ibpf} and \code{bpfilter}.
+
+The success of an iterated filter for likelihood maximization depends on the success of a single pass of that filter for likelihood evaluation.
+A useful diagnostic quantity for the particle filter is the effective sample size (ESS) \citep{douce11,liu01}.
+In the notation of \citep{king16}, the filter observation weight for particle $j$ at time $n$ is $w(n,j)=f_{Y_n|X_n}(y_n^*|X^P_{n,j};\theta)$, the normal weight is $\tilde{w}(n,j)=w(n,j)\big/ \sum_{k=1}^Jw(n,k)$, and the effective sample size is
+\begin{equation}
+\mathrm{ESS}_n = \left(\sum_{j=1}^J \tilde{w}(n,j)^2\right)^{-1}.
+\end{equation}
+ESS is generally motivated as an approximation to the equivalent number of independent samples from the filtering distribution. 
+In the context of likelihood-based inference, it is convenient to think of $\mathrm{ESS}^{-1}_n$ as an approximation to the variance of the conditional log-likelihood estimate,
+\begin{equation}
+\hat \ell_n = \log \left( \frac{1}{J} \sum_{j=1}^J w(n,j) \right).
+\end{equation}
+However, ubiquitous multi-core computation permits us to calculate this variance directly, by Monte Carlo replication.
+This bypasses consideration of when ESS is, and is not, a good estimator. 
+Also, the $R$ Monte Carlo replicates, $\hat\ell_n^{(r)}$, $r=1,\dots,R$, give rise to an improved estimate of the conditional log-likelihood,
+\begin{equation}
+\bar \ell_n = \frac{1}{R} \sum_{r=1}^R \hat\ell_n^{(r)},
+\end{equation}
+together with an uncertainty estimate via the central limit theorem.
+
+<<pfilter-test-london,eval=T,echo=F,fig.height=6, fig.width=6, out.width="4.5in", fig.cap = "Comparison of effective sample size and conditional log-likelihood variance for measles in London">>=
+london <- he10(U=1)
+london_diag <- bake(file=paste0(diagnostics_dir,"london.rds"),seed=230604, {
+  pf_list <- foreach(rep=1:switch(i,6,100)) %dopar% pfilter(london,Np=switch(i,10,5000))
+  ess=sapply(pf_list,function(x)x@eff.sample.size)
+  ll=sapply(pf_list,function(x)x@cond.logLik)
+  list(mess=apply(ess,1,mean),cllv=apply(ll,1,var),mll=apply(ll,1,mean))
+})
+london_ess <- london_diag$mess; london_cllv <- london_diag$cllv; londom_mll <- london_diag$mll
+xy_range <- range(c(1/london_ess,london_cllv))
+plot(x=london_cllv,y=1/london_ess,log="xy",ylab="1/ESS",xlab="Conditional log-likelihood variance",
+  xlim=xy_range, ylim = xy_range)
+abline(a=0,b=1)
+@
+
+Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
+We see that both measures agree on the most problematic observation.
+However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution, and so we prefer to work with CLLV.
+CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
+Using the notation of \citep{asfaw23}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
+\begin{equation}
+w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
+            f_{Y_{n,n}|X_{u,n}}
+            \big(
+              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
+            \big).
+\end{equation}
+The BCLL is
+\begin{equation}
+\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
+\end{equation}
+and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
+\begin{equation}
+\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
+\end{equation}
+For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
+<<bpfilter-diag-code,eval=F,echo=T>>=
+bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
+  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
+}
+bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
+dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
+bcll <- apply(bpf_ll,c(1,2),mean)
+bcllv <- apply(bpf_ll,c(1,2),var)
+@
+
+Once we have identified that an observation is hard to filter, there are various classes of explanation: (i) the data point is an outlier, which any reasonable model would struggle to explain; (ii) the model is poor at that point; (iii) there is no problem with the model or data, but the system happens to require a high Monte Carlo effort at that point.
+Neither ESS nor CLLV can distinguish these alternatives.
+It can be helpful to compare the model fit against a simple statistical model, to provide an objective benchmark.
+Measles case counts are an example of a population system exhibiting exponential growth and decay, and in such cases a log-scale autoregressive moving average (log-ARMA) model can provide an appropriate benchmark.
+The function \code{spatPomp::arma\_benchmark} fits this model for each unit of a \code{spatPomp} object and provides the unit conditional log-likelihood values for each time point.
+Here, blocks are taken to be units, so we define the log-likelihood anomaly as the difference between the BCCL and the unit benchmark,
+<<anomaly,echo=T,eval=F>>=
+benchmark <- arma_benchmark(he10_model)
+anomaly <- bcll - benchmark$cond
+@
+Unlike BCLL, both log-likelihood anomaly and BCLLV have the convenient property that they do not depend on the scale of the data, or the units in which it is measured.
+
+Figure~\ref{fig:bpfilter-diagnostics} investigates two units---London and Birmingham---via BCLLV and log-likelihood anomaly.
+We see that the model used here fits London well and Birmingham poorly.
+Some time points have very high Monte Carlo likelihood estimate variance, and those points also have poor likelihood compared to the benchmark.
+One observation for London has a poor log-likelihood anomaly but a decent BCLLV.
+For Birmingham, both anomaly and BCLLV lead the same conclusion that the model is unsatisfactory, at these default initial parameter values.
+Here, our goal is only to check methods on simulated data, but if we wanted to fit the model to data, we should confirm whether the diagnostics from the final fitted model have improved.
+
+<<bpfilter-diagnostics,eval=T,echo=F,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Block particle filter diagnostics for London (A,C) and Birmingham (B,D).">>=
+bpf_diag <- bake(file=paste0(diagnostics_dir,"bpfilter.rds"),seed=230604, {
+<<bpfilter-diag-code>>
+  list(bcll=bcll,bcllv=bcllv)
+})
+bcll <- bpf_diag$bcll; bcllv <- bpf_diag$bcllv
+<<anomaly>>
+par(mfrow=c(2,2))
+par(mai=c(1,0.8,0.1,0.1))
+plot(x=bcllv[1,],y=anomaly[1,],log="x",ylab="anomaly",xlab="Conditional log-likelihood variance")
+mtext("A",side=3,line=-1,adj=-0.24,cex=1.5)
+plot(x=bcllv[2,],y=anomaly[2,],log="x",ylab="anomaly",xlab="Conditional log-likelihood variance")
+mtext("B",side=3,line=-1,adj=-0.24,cex=1.5)
+plot(y=anomaly[1,],x=time(he10_model),ylab="anomaly",xlab="Date")
+mtext("C",side=3,line=-1,adj=-0.24,cex=1.5)
+plot(y=anomaly[2,],x=time(he10_model),ylab="anomaly",xlab="Date")
+mtext("D",side=3,line=-1,adj=-0.24,cex=1.5)
+@
+
+Trace plots to check on the convergence of \code{ibpf} are similar to those for other iterated filtering algorithms.
+In Figure~\ref{fig:traces}, we show this for the Gaussian correlated random walk example.
+In this case, we see replicable convergence among independent searches.
+
+<<traces,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Trace plots for the correlated Gaussian random walk example.">>=
+ibpf_traces <- pomp::melt(lapply(ibpf_mle_searches,pomp:::traces_internal))
+ibpf_traces$iteration <- as.numeric(ibpf_traces$iteration)
+ggplot(ibpf_traces, aes(x=iteration,y=value,group=.L1,color=factor(.L1))) +
+  geom_line() +
+  guides(color="none") +
+  facet_wrap(~variable,scales="free_y")
+@
+
 \bibliography{bib-tutorial}

### 2023-06-08 ms.Rnw
tidying new diagnostics section

@@
 
+\clearpage
+
 \section{Diagnostics}
@@
 Even if inference appears to be operating satisfactorily, it is good practice to see whether there are some data points for which the model is inappropriate, or some parameters for which optimization is less stable.
-It may be interesting to know if less Monte Carlo effort would be sufficient, or if more would be useful.
+It may be interesting to know if less Monte Carlo effort would be sufficient, or if greater effort would be beneficial.
 For all these questions, a starting point is exploratory analysis of diagnostic data produced by \code{ibpf} and \code{bpfilter}.
@@
 The success of an iterated filter for likelihood maximization depends on the success of a single pass of that filter for likelihood evaluation.
-A useful diagnostic quantity for the particle filter is the effective sample size (ESS) \citep{douce11,liu01}.
-In the notation of \citep{king16}, the filter observation weight for particle $j$ at time $n$ is $w(n,j)=f_{Y_n|X_n}(y_n^*|X^P_{n,j};\theta)$, the normal weight is $\tilde{w}(n,j)=w(n,j)\big/ \sum_{k=1}^Jw(n,k)$, and the effective sample size is
+A useful diagnostic quantity for the particle filter is the effective sample size (ESS) \citep{doucet11,liu01}.
+In the notation of \citet{king16}, the filter observation weight for particle $j$ at time $n$ is $w(n,j)=f_{Y_n|X_n}(y_n^*|X^P_{n,j};\theta)$, the normal weight is $\tilde{w}(n,j)=w(n,j)\big/ \sum_{k=1}^Jw(n,k)$, and the effective sample size is
 \begin{equation}
@@
 However, ubiquitous multi-core computation permits us to calculate this variance directly, by Monte Carlo replication.
-This bypasses consideration of when ESS is, and is not, a good estimator. 
-Also, the $R$ Monte Carlo replicates, $\hat\ell_n^{(r)}$, $r=1,\dots,R$, give rise to an improved estimate of the conditional log-likelihood,
+This bypasses consideration of whether or not ESS is a good estimator. 
+Also, the Monte Carlo replicates, denoted by $\hat\ell_n^{(r)}$, for $r=1,\dots,R$, give rise to an improved estimate of the conditional log-likelihood,
 \begin{equation}
+\label{eq:ell-bar}
 \bar \ell_n = \frac{1}{R} \sum_{r=1}^R \hat\ell_n^{(r)},
@@
 together with an uncertainty estimate via the central limit theorem.
+One could also replace the arithmetic average  in (\ref{eq:ell-bar}) using \code{pomp::logmeanexp}.
 
@@
 We see that both measures agree on the most problematic observation.
-However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution, and so we prefer to work with CLLV.
+However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution.
 CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
-Using the notation of \citep{asfaw23}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
+Using the notation of \citet{asfaw23arxiv}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
 \begin{equation}
@@
 
+Log-likelihood anomalies can be plotted against any covariate of interest for exploratory data analysis.
+This is analogous to residual analysis for linear models.
+
 Trace plots to check on the convergence of \code{ibpf} are similar to those for other iterated filtering algorithms.

### 2024-08-05 ms.Rnw
some updates

@@
 <<bm-followup,echo=F,eval=T>>=
+if(i==1) cores <- 2   # no need to use all available cores when i=1
 U <- length(unit_names(b))
@@
 m_g <- simulate(he10_model,seed=27)
-if(dim(obs(m_g))!=dim(obs(m_R0)) || any(obs(m_g)-obs(m_R0) != 0)) stop("m_g and m_R0 should have the same synthetic data")
+if(any(dim(obs(m_g))!=dim(obs(m_R0))) || any(obs(m_g)-obs(m_R0) != 0)) stop("m_g and m_R0 should have the same synthetic data")
 })

### 2024-08-06 ms.Rnw
re-run on greatlakes
recompiled and update references

@@
 \author{Ning Ning and Edward L. Ionides}
-\date{\small Compiled \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}\\
+\date{\small This version compiled on \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}\\
 Source code at \url{https://github.com/ionides/ibpf-tutorial}}
@@
 
+\section{Discussion}
+
+Further examples of \code{ibpf} being used for data analysis are provided by \citet{li24} and \citet{wheeler24}.
+\citet{li24} studied the spread of COVID-19 in China at the start of the pandemic.
+Their models are available in the \code{metapoppkg} R package (\url{https://doi.org/10.5281/zenodo.10149233}) and the article source code is available at \url{https://doi.org/10.5281/zenodo.10149258}.
+\citet{wheeler24} studied a cholera epidemic in Haiti.
+Their models are available in the \code{haitipkg} R package (\url{https://doi.org/10.5281/zenodo.7557099}) and the article source code is available at \url{https://doi.org/10.5281/zenodo.10783080}.
+
 \bibliography{bib-tutorial}

### 2024-08-06 ms.Rnw
edits based on Lees review - adding purled script

@@
 \renewcommand\Yspace{\R}
-%\newcommand\Thetadim{{\mathrm{dim}}(\Thetaspace)}
 \newcommand\Thetadim{D_\theta}
 \newcommand\thetadim{d_\theta}
-%\newcommand\vectheta{\vec{\theta}}
 \newcommand\vectheta{\theta}
@@
 \newcommand\IVP{\hspace{0.15mm}\mathrm{IVP}}
-
-%% for bagged filters
-\newcommand\ABF{ABF}
-\newcommand\ABFIR{ABF-IR}
-\newcommand\UBF{UBF}
-\newcommand\rep{i}
-\newcommand\altRep{\tilde\rep}
-\newcommand\Rep{\mathcal{I}}
 
@@
 \newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
 \newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
@@
 \newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
 \newcommand\bigO{\mathcal{O}}
@@
 library("knitr")
-stopifnot(packageVersion("pomp")>="4.1")
+stopifnot(packageVersion("pomp")>="5.0")
 @
@@
     progress=TRUE,prompt=TRUE,highlight=FALSE,
-#    tidy=TRUE,
-#    tidy.opts=list(
-#        keep.blank.line=FALSE
-#    ),
     comment="",
@@
 \date{\small This version compiled on \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}\\
-Source code at \url{https://github.com/ionides/ibpf-tutorial}}
+\vspace{3mm}
+Source code at \url{https://github.com/ionides/ibpf-tutorial}\\
+\vspace{3mm}
+R script at \url{https://github.com/ionides/ibpf-tutorial/blob/main/ms.R}}
+
 \begin{document}
@@
 We also test \code{ibpf} on simulated data for a measles transmission model.
-In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file. 
 \end{abstract}
@@
 if(i==1) cores <- 2   # no need to use all available cores when i=1
-U <- length(unit_names(b))
 @
@@
 For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
-This study uses a sufficiently small number of units (\code{U=\Sexpr{U}}) that the particle filter is numerically tractable.
+This study uses a small number of units,
+<<define-U,echo=T>>=
+U <- 4
+@
+so that the particle filter is numerically tractable.
 We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.
@@
 
-<<bm_logLik_code,cache=F,echo=T,eval=F>>=
+<<bm_logLik_code,cache=F,echo=T,eval=F,purl=F>>=
 kf_logLik <- bm2_kalman_logLik(b)
@@
 \citet{ionides21} investigated a range of values for $U$ for this model in their Figure~1, and for an epidemiological model their Figure~3.
-The small scenario considered here, with $U=4$, is designed for the following purposes: (i) to validate whether or not \code{ibpf} is correctly coded by comparison with direct calculations using the Kalman filter; (ii) to check whether or not the block approximation has considerable adverse effects on inference in this case.
-The inherent scalability of BPF and IBPF, together with the spatial homogeneity of the model, suggests that results for $U=4$ should be representative for larger $U$.
+The small scenario considered here, with $U=\Sexpr{U}$, is designed for the following purposes: (i) to validate whether or not \code{ibpf} is correctly coded by comparison with direct calculations using the Kalman filter; (ii) to check whether or not the block approximation has considerable adverse effects on inference in this case.
+The inherent scalability of BPF and IBPF, together with the spatial homogeneity of the model, suggests that results for $U=\Sexpr{U}$ should be representative for larger $U$.
 
@@
 names(prof_init) <- paste0("rho",1:U)
-prof_vals <- seq(from=0,to=0.8,length=switch(i,5,20))
+prof_vals <- seq(from=0,to=0.8,length.out=switch(i,5,20))
 
@@
 
-<<ibpf-mle-code,eval=F,echo=T>>=
-rho_start <- seq(from=0.2,to=0.8,length=U)
+<<ibpf-mle-code,eval=F,echo=T,purl=F>>=
+rho_start <- seq(from=0.2,to=0.8,length.out=U)
 params_start <- coef(b)
@@
   u <- 1
-  rho_start <- seq(from=0.2,to=0.8,length=U)
+  rho_start <- seq(from=0.2,to=0.8,length.out=U)
   reps <- switch(i,3,10)
@@
 names(prof_sig_init) <- paste0("sigma",1:U)
-prof_sig_vals <- seq(from=0.4,to=1.6,length=switch(i,5,20))
+prof_sig_vals <- seq(from=0.4,to=1.6,length.out=switch(i,5,20))
 
@@
 u <- 1
-sigma_start <- seq(from=0.4,to=1.6,length=U)
+sigma_start <- seq(from=0.4,to=1.6,length.out=U)
 params_sig_start <- coef(b_sig)
@@
 
-<<model-he10-code,eval=F,echo=T>>=
+<<model-he10-code,eval=F,echo=T,purl=F>>=
 he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1955,1964),
@@
 stew(file=paste0(m_dir,"m_prof_R02.rda"),seed=722,{
-m_prof_R0_vals <- seq(from=25,to=35,length=switch(i,5,20))
+m_prof_R0_vals <- seq(from=25,to=35,length.out=switch(i,5,20))
 tic <- Sys.time()
@@
 slice_points <- switch(i,5,20)
-slice_g_vals <- rep(seq(from=100,to=2500,length=slice_points),each=slice_reps)
+slice_g_vals <- rep(seq(from=100,to=2500,length.out=slice_points),each=slice_reps)
 params_g_slice<- matrix(coef(m_g),
@@
 For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
-<<bpfilter-diag-code,eval=F,echo=T>>=
+<<bpfilter-diag-code,eval=F,echo=T,purl=F>>=
 bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {

### 2024-08-06 ms.R

@@ -1,0 +1,686 @@
+## ----cores,echo=F,cache=F,message = FALSE, warnings = FALSE,results='hide'----
+library(doParallel)
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+ggplot2::theme_set(ggplot2::theme_bw())
+
+
+## ----packages,include=F,echo=F,cache=F----------------------------------------
+library("spatPomp")
+library("ggplot2")
+library("tidyverse")
+library("knitr")
+stopifnot(packageVersion("pomp")>="5.0")
+
+
+
+
+
+
+## ----setup,cache=F,include=F--------------------------------------------------
+myround<- function (x, digits = 1) {
+  # adapted from the broman package
+  # solves the bug that round() kills significant trailing zeros
+  if (length(digits) > 1) {
+    digits <- digits[1]
+    warning("Using only digits[1]")
+  }
+  if (digits < 1) {
+    as.character(round(x,digits))
+  } else {
+    tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
+    zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
+    tmp[tmp == paste0("-", zero)] <- zero
+    tmp
+  }
+}
+mysignif <- function (x, digits = 1) {
+  myround(x, digits - ceiling(log10(abs(x))))
+}
+
+
+## ----bm-setup,echo=F,eval=T---------------------------------------------------
+set.seed(20)
+
+
+## ----bm-model,echo=T,eval=T---------------------------------------------------
+library(spatPomp)
+i <- 1
+b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
+plot(b)
+
+
+## ----bm-followup,echo=F,eval=T------------------------------------------------
+if(i==1) cores <- 2   # no need to use all available cores when i=1
+
+
[…]

### 2024-08-07 ms.Rnw
a few more minor edits

@@
 i <- 2
-b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
+U <- 4
+b <- bm2(U=U,N=switch(i,10,200),unit_specific_names="rho")
 plot(b)
@@
 For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
-This study uses a small number of units,
-<<define-U,echo=T>>=
-U <- 4
-@
-so that the particle filter is numerically tractable.
+This study uses a small number of units ($U=\Sexpr{U}$) so that the particle filter is numerically tractable.
 We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.

### 2024-08-07 ms.R

@@ -46,7 +46,7 @@ set.seed(20)
 
 ## ----bm-model,echo=T,eval=T---------------------------------------------------
 library(spatPomp)
-i <- 1
+i <- 2
 b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
 plot(b)
 

### 2024-08-07 ms.R
a typo

@@ -47,16 +47,13 @@ set.seed(20)
 ## ----bm-model,echo=T,eval=T---------------------------------------------------
 library(spatPomp)
 i <- 2
-b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
+U <- 4
+b <- bm2(U=U,N=switch(i,10,200),unit_specific_names="rho")
 plot(b)
 
 
 ## ----bm-followup,echo=F,eval=T------------------------------------------------
 if(i==1) cores <- 2   # no need to use all available cores when i=1
-
-
-## ----define-U,echo=T----------------------------------------------------------
-U <- 4
 
 
 

### 2024-10-25 ms.Rnw
ben bolker suggestions incorporated

@@
         stringsAsFactors=FALSE,
-        prompt="R> ",
-        continue="+  ",
+        prompt="   ",
+        continue="   ",
         width=70,
@@
 For this model, exact likelihood evaluation is not available.
-However, for a relatively small number of units ($U=4$) the particle filter provide an adequate approximation.
+However, for a relatively small number of units ($U=4$) the particle filter provides an adequate approximation.
 \citet{ionides22} considered fitting this model to data using IBPF, with 20 cities and up to $20\times 13$ parameters.
@@
 We choose two large towns (London and Birmingham) and two small towns (Cardiff and Hastings) since we expect that population movement from large towns to small towns is essential to explain disase persistence in small towns.
-Large towns can maintain an ongoing epidemic, but, below a critical community size local extinction of the disease is expected during epidemic troughs.
+Large towns can maintain an ongoing epidemic, but below a critical community size local extinction of the disease is expected during epidemic troughs.
 
@@
 Likelihood evaluation took \Sexpr{myround(as.double(m_lik_eval_time,units="mins"),2)} mins.
-Recall that the bias-variance trade-off for likelihood evaluation becomes a tradeoff between two sources of bias for log-likelihood evaluation, due to Jensen's inqeuality.
+Recall that the bias-variance trade-off for likelihood evaluation becomes a tradeoff between two sources of bias for log-likelihood evaluation, due to Jensen's inquality.
 Table~\ref{tab:m-lik-eval} shows that little likelihood is lost due to the block approximation for small numbers of units.
 Indeed, even for a large number of particles, a small block size gives higher log-likelihood estimates.
-This is on contrast to the results in Table~\ref{tab:lik-eval} for the spatially correlated random walk example in Section~\ref{sec:bm}.
+This is in contrast to the results in Table~\ref{tab:lik-eval} for the spatially correlated random walk example in Section~\ref{sec:bm}.
 
@@
 If IBPF is effectively maximizing the BPF approximation to the likelihood, then situations where BPF has low likelihood evaluation bias may correspond to situations where IBPF has low estimation bias.
-A profile likelihood for one paramter is presented to support this, in Figure~\ref{fig:m-prof-R0-plot}.
+A profile likelihood for one parameter is presented to support this, in Figure~\ref{fig:m-prof-R0-plot}.
 Here, the evaluation using PF gives a slightly tighter estimate of the profile, but this may be less accurate: PF is a higher variance algorithm, even with $U=4$, and its variance increases as the model becomes increasingly misspecified.
@@
 Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
-We see that both measures agree on the most problematic observation.
+We see that both measures agree on the most problematic observations.
 However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution.
@@
 \end{equation}
-For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
+For our measles spatPomp example, we calculate the BCLL and its variance (BCLLV) as follows:
 <<bpfilter-diag-code,eval=F,echo=T,purl=F>>=
@@
 In this case, we see replicable convergence among independent searches.
-
-<<traces,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Trace plots for the correlated Gaussian random walk example.">>=
+The likelihood rises toward the maximum and then stabilizes in a neighborhood of the maximum while the estimated parameters explore this region.
+
+<<traces,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Trace plots for the correlated Gaussian random walk example.
+Constant traces for $\\sigma_1$, $\\tau_1$ and $X_{0,1}$ arise because these parameters are treated as known and not estimated.
+Note that the values of $\\sigma_1$, $\\tau_1$ and $X_{0,1}$  are shared across units (e.g., there is no $\\sigma_2$) since the \\code{bm2} constructor is called with $\\rho$ as the only unit-specific parameter.">>=
 ibpf_traces <- pomp::melt(lapply(ibpf_mle_searches,pomp:::traces_internal))
@@
 
+\section*{Acknowledgments}
+
+We are grateful to Ben Bolker for providing useful feedback.
+
 \bibliography{bib-tutorial}
